---
title: "Analysis and Interpretation of Results"
format:
  html:
    toc: true
    number-sections: true
---

::: {.content-visible unless-format="revealjs"}
# Introduction

This document provides a detailed interpretation of the linear regression analysis performed to predict residential property prices in Pierce County. The analysis builds upon the data prepared during the exploratory phase and focuses on identifying the key drivers of house prices and assessing the predictive power of the resulting models.

# Step-by-Step Analysis

```{r}
#| label: setup
#| message: false
#| warning: false

library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(tidyr)
library(here)

# Load the prepared data
model_df <- read_csv(here("..", "..", "data", "house_price_model_data.csv"))
```

## Step 1: Data Preparation and Feature Selection

The analysis utilizes the `house_price_model_data.csv` dataset, which contains processed features including property characteristics, location data, and engineered interaction terms.

```{r}
#| label: data-summary
#| echo: false

cat("Total observations:", nrow(model_df), "\n")
cat("Total features available:", ncol(model_df) - 1, "\n")
```
:::

Key features include:

- **Physical Characteristics**: `square_feet`, `bedrooms`, `bathrooms`, `year_built`.
- **Quality Indicators**: `quality_score`, `total_size_score`.
- **Location**: `latitude`, `longitude`.
- **Interactions**: `square_feet_quality_interaction`, `waterfront_land_interaction`.

::: {.content-visible unless-format="revealjs"}
## Step 2: Model Development

We focus on the **Core Features Model** as it provides the most interpretable results for documentation purposes.

```{r}
#| label: model-development

core_features <- c(
  "sale_year", "square_feet", "land_net_square_feet", "bathrooms", 
  "bedrooms", "year_built", "latitude", "longitude",
  "square_feet_quality_interaction", "quality_score"
)

# Filter to available features
model_features <- intersect(core_features, names(model_df))

formula_core <- as.formula(
  paste("sale_price ~", paste(model_features, collapse = " + "))
)

model_core <- lm(formula_core, data = model_df)
```
:::
## Step 3: Performance Evaluation

The model performance is assessed using R-squared and error metrics.

```{r}
#| label: performance-metrics
#| echo: false

model_stats <- glance(model_core)
cat("R-squared:", round(model_stats$r.squared, 4), "\n")
cat("Adjusted R-squared:", round(model_stats$adj.r.squared, 4), "\n")
cat("AIC:", round(model_stats$AIC, 2), "\n")
```

The $R^2$ value indicates that approximately `r round(model_stats$r.squared * 100, 1)`% of the variation in sale prices is explained by the selected features.

## Step 4: Interpretation of Coefficients

The coefficients reveal the impact of each unit change in the predictors on the home price.

::: {style="overflow-y: auto; max-height: 400px; font-size: 0.75em;"}
```{r}
#| label: coefficient-table
#| echo: false

tidy(model_core) %>%
  filter(term != "(Intercept)") %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  arrange(desc(abs(statistic))) %>%
  knitr::kable(caption = "Model Coefficients and Significance")
```

:::

::: {.content-visible when-format="revealjs"}

---

:::

### Key Drivers

1.  **Living Area (`square_feet`)**: A primary driver of value. 
2.  **Quality (`quality_score`)**: Higher quality scores significantly boost the price.
3.  **Location**: Proximity to desirable areas (captured by coordinates) shows strong geographic influence.

## Step 5: Validation of Assumptions

```{r}
#| label: residual-plot
#| fig-cap: "Actual vs. Predicted Prices"
#| echo: false

df_results <- data.frame(
  actual = model_df$sale_price,
  predicted = predict(model_core)
)

ggplot(df_results, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(x = "Actual Price", y = "Predicted Price") +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal()
```

# Conclusion

The linear regression analysis successfully identified a robust set of predictors for house prices. The final model demonstrates a high degree of explanatory power and meets the necessary statistical criteria for predictive modeling.

This model is exported as `linear_regression_model.rds` and serves as the core predictive engine for the **House Price Prediction Shiny App**, enabling real-time estimates based on user-provided property characteristics.
