---
format:
  html:
    toc: true
    number-sections: true
  revealjs: default
---

# KNN model

::: {.content-visible unless-format="revealjs"}
## Introduction

This document provides a detailed interpretation of the KNN model applied to predict residential property prices in Pierce County. The analysis builds upon the data prepared during the exploratory phase and focuses on identifying the key drivers of house prices and assessing the predictive power of the resulting models.

## Step-by-Step Analysis

```{r}
#| label: setup_2
#| message: false
#| warning: false

library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(tidyr)
library(here)
library(dotenv)
library(caret)
library(tidyverse)
library(kknn)
options(scipen = 999)

# Load the prepared data
model_df <- read.table(here::here("..", "..", "data", "app_model_df.txt"), header = TRUE, sep = "\t")
```


### Step 1: Data Preparation and Feature Selection

The analysis utilizes the `house_price_model_data.csv` dataset, which contains processed features including property characteristics and location data.

```{r}
#| label: data-summary-2
#| echo: false

cat("Total observations:", nrow(model_df), "\n")
cat("Total features available:", ncol(model_df) - 1, "\n")
```

The idea behind experimenting with KNN is to easily show the user why the given price
has been estimated through a clear reasoning (showing n most similar observations)
based on a very limited number of features (location and size, optionally condition).

Key features include:

- **Physical Characteristics**: `square_feet`, `bathrooms`, `bedrooms`
- **Quality Indicators**: `quality_score`, 
- **Location**: `latitude`, `longitude`
- **Time**: `sale year`.

:::

### Step 2: Model Development

::: {.content-visible unless-format="revealjs"}
We developed four KNN models iteratively, adding features at each step to explore 
how different dimensions of the data contribute to price prediction.
:::

::: {style="font-size: 0.75em;"}

The **baseline model** uses the three core spatial and size features — latitude, 
longitude, and square footage — as a minimal starting point.

In the **second model**, we added `sale_year`, motivated by its correlation with 
sale price, to capture the temporal dimension of the market.

The **third model** incorporates `quality_score`, introducing a condition and 
quality dimension that reflects the physical state of the property beyond its size alone.

Finally, the **full model** adds `bathrooms` and `bedrooms`, rounding out the 
feature set with the most common descriptors of residential property layout.

:::

### Step 3: Model Evaluation
::: {style="overflow-y: auto; max-height: 400px; font-size: 0.75em;"}
```{r}
# load (in any other chunk/file)
metrics_table   <- readRDS(here("..", "..","analysis", "03-knn-files", "metrics_table.rds"))
within_15_table <- readRDS(here("..", "..","analysis", "03-knn-files", "within_15_table.rds"))
```

| Model | k | RMSE | R-squared | MAE |
|:----------|:-------|:-------|:-------|:-------|
| **minimal** | 5 | `r metrics_table[1, "RMSE"]` | `r metrics_table[1, "Rsquared"]` | `r metrics_table[1, "MAE"]` |
| **+sale_year** | 2 | `r metrics_table[2, "RMSE"]` | `r metrics_table[2, "Rsquared"]` | `r metrics_table[2, "MAE"]` |
| **+sale_year+quality** | 2 | `r metrics_table[3, "RMSE"]` | `r metrics_table[3, "Rsquared"]` | `r metrics_table[3, "MAE"]` |
| **full** | 2 | `r metrics_table[4, "RMSE"]` | `r metrics_table[4, "Rsquared"]` | `r metrics_table[4, "MAE"]` |

| Model | within 15% train | within 15% test |
|:----------|:-------|:-------|
| **minimal** | `r within_15_table[1, "within_15_train"]` | `r within_15_table[1, "within_15_test"]` |
| **+sale_year** | `r within_15_table[2, "within_15_train"]` | `r within_15_table[2, "within_15_test"]` |
| **+sale_year+quality** | `r within_15_table[3, "within_15_train"]` | `r within_15_table[3, "within_15_test"]` |
| **full** | `r within_15_table[4, "within_15_train"]` | `r within_15_table[4, "within_15_test"]` |

:::

