# Variable selection and Data Cleaning

After all the data has been mapped put in a database we construct the final selection
of variables used for modelling and clean this data.

```{r}

library(DBI)
library(RPostgres)
library(dotenv)
library(here)
library(dplyr)
library(tidyr)
library(forcats)
library(stringr)

load_dot_env(here::here(".env"))
```




```{r}
con <- dbConnect(
  RPostgres::Postgres(),
  host     = Sys.getenv("DB_HOST"),
  port     = as.integer(Sys.getenv("DB_PORT")),
  dbname   = Sys.getenv("DB_NAME"),
  user     = Sys.getenv("DB_USER"),
  password = Sys.getenv("DB_PASSWORD")
)
dbIsValid(con)
```


## Subset the Data

After mapping the data out and looking at the variable description, we limit the
potentially used features. These features should be ones, that a homeowner likely
knows about their house.

### Land and Location Attributes

Restrict the features to:
- land square feet
- waterfront type
- view quality
- electricity availability
- water availability
- street type
- coordinates (longitude, latitude)

```{r}
#| label: tbl-appraisal-features
#| tbl-cap: "Selected appraisal account features: land and location attributes"

df_1 <- dbGetQuery(con, '
  SELECT parcel_no, appraisal_account_type, buildings, land_net_squarefeet, waterfront_type, view_quality,
  utility_electric, utility_water, street_type, latitude, longitude
  FROM "APPRAISAL_ACCOUNT"
  WHERE appraisal_account_type = \'Residential\'
')
```

### Property Characteristics

We get property specifics:
- condition
- number of stories
- number of bedrooms
- number of bathrooms
- units
- year built
- year remodeled

```{r}
#| label: tbl-improvement-features
#| tbl-cap: "Selected improvement features: building characteristics and construction details"

df_2 <- dbGetQuery(con, "
  SELECT i.parcel_no, i.building_id, i.property_type, i.condition,
         b.stories, b.bedrooms, b.bathrooms, b.units, b.year_built, b.year_remodeled
  FROM \"IMPROVEMENT\" i
  LEFT JOIN \"IMPROVEMENT_BUILT_AS\" b
    ON i.parcel_no = b.parcel_no
    AND i.building_id = b.building_id
  WHERE i.property_type = 'Residential'
")
```

### Property Valuation

We get the current property value:

```{r}
#| label: tbl-valuation-feature
#| tbl-cap: "Market value from tax account table (target variable)"

df_3 <- dbGetQuery(con, '
  SELECT parcel_no, total_market_value_current_year
  FROM "TAX_ACCOUNT"
')
```

### Data Integration

We join the data together from all three source subsets:

```{r}
#| label: tbl-integrated-dataset
#| tbl-cap: "Integrated dataset combining land, improvement, and valuation features"

df_all <- df_1 %>%
  left_join(df_2, by = "parcel_no") %>%
  left_join(df_3, by = "parcel_no")
```


### Cleaning

We obtain the dataset with selected features. Now we go ahead to prepare the data
for initial exploratory data analysis.

```{r}
df_all <- df_all %>% 
  select( -property_type)
```

We first select lands with buildings on it, as we are interested in residential
home valuations.
```{r}
df_all <- df_all %>% 
  filter(buildings != 0)
```

Then we look at the missing values and treat them one-by-one.

```{r}
#| label: tbl-missing-values-initial
#| tbl-cap: "Initial assessment of missing values across all features"

colSums(is.na(df_all))
```

We remove observations that have no bedroom or bathroom recorded as these would
be one of the features easiest for the user to input and we want to use them. 
If they are missing it likely hints at a data recording issue or a building type
that is not a classical residential home (which is out of scope for this project).
```{r}
df_all <- df_all %>% 
  filter(!is.na(bathrooms))
```

We also remove the 5 observations missing coordinates or square footage.
```{r}
df_all <- df_all %>% 
  filter(!is.na(latitude) & !is.na(longitude) & !is.na(land_net_squarefeet))
```

Waterfront type records River, Lake, Stream/Creek or Salt. We fill in missing
values to represent no waterfront, as such category does not exist and missing
specification likely represents that the property ahs no direct water connection.

```{r}
df_all <- df_all %>% 
  mutate(waterfront_type = replace_na(waterfront_type, "None"))
unique(df_all$waterfront_type)
```

View quality currently has 9 categories that is slightly excessive, and more confusing for
the potential user and increases model complexity. 
```{r}
unique(df_all$view_quality)
```

We collapse the categories into "limited", "average", "good", "very good". We replace
NAs with uncategorized.

```{r}
df_all <- df_all %>%
  mutate(
    view_quality = replace_na(view_quality, "uncategorized"),

    view_quality = fct_collapse(view_quality,
      "limited"   = c("View Lim", "View Lim -", "View Lim +"),
      "average"   = c("View Avg", "View Avg +"),
      "good"      = c("View Good", "View Good +"),
      "very good" = c("View V-Good", "View V-Good +")
    )
  )
unique(df_all$view_quality)
```

We order the view types.
```{r}
df_all$view_quality <- factor(df_all$view_quality, 
                               levels = c("uncategorized", "limited", "average", "good", "very good"),
                               ordered = TRUE)
```

For the year remodeled, we will proceed as no remodeling happened, if the year is
not recorded.
```{r}
df_all <- df_all %>% 
  mutate(year_remodeled = replace_na(year_remodeled, 0))
```

Lastly, in the units feature we convert the few missing values to 1 unit after looking
into the data manually, and also convert 0 units to 1 as the empirical meaning
of having no separate units and the whole unit being one (for example family home)
is the same.
```{r}
df_all <- df_all %>%
  mutate(units = if_else(is.na(units) | units == 0, 1, units))
```

We have now the complete dataset with the preselected features.

```{r}
#| label: tbl-missing-values-final
#| tbl-cap: "Final verification of missing values after cleaning (all zeros indicate complete data)"

colSums(is.na(df_all))
```

### Data Quality Validation

A few post checks we should still apply.

```{r}
unique(df_all$utility_electric)
df_all <- df_all %>%
  mutate(utility_electric = str_replace(utility_electric, "POWER NO - COMMENT", "POWER NO"))
unique(df_all$utility_electric)
```
```{r}
unique(df_all$utility_water)
```
```{r}
unique(df_all$street_type)
```

Check coordinate range.

```{r}
#| label: tbl-coordinate-range
#| tbl-cap: "Geographic bounding box of properties in the dataset"

df_all %>%
  summarise(
    min_lat = min(latitude, na.rm = TRUE),
    max_lat = max(latitude, na.rm = TRUE),
    min_lon = min(longitude, na.rm = TRUE),
    max_lon = max(longitude, na.rm = TRUE)
  )
```
```{r}
unique(df_all$condition)
df_all$condition <- factor(df_all$condition, 
                                levels = c("Uninhabitable", 
                                           "Extra Poor", 
                                           "Very Poor", 
                                           "Poor", 
                                           "Fair", 
                                           "Average"),
                                ordered = TRUE)
```

```{r}
df_all %>%
  select(stories, bedrooms, bathrooms) %>%  
  slice_max(order_by = stories, n = 5)
```
The top 3 observation are very unlikely with this combination of bedrooms and bathrooms,
so we remove them.
```{r}
to_remove <- df_all %>%
  slice_max(order_by = stories, n = 3, with_ties = FALSE)

df_all <- df_all %>%
  anti_join(to_remove)
```
Looking at the data again manually, and in particular the combination of bathrooms
and bedrooms with teh units we also detect some anomalies that we remove.
```{r}
df_all <- df_all %>%
  filter(bathrooms <= 17 & bedrooms <= 21)
```

The year remodeled being 0 means no remodeling has been recorded on the house.

```{r}
#| label: tbl-year-range
#| tbl-cap: "Range of construction and renovation years (0 = no remodeling recorded)"

df_all %>%
  summarise(
    min_built     = min(year_built, na.rm = TRUE),
    max_built     = max(year_built, na.rm = TRUE),
    min_remodeled = min(year_remodeled, na.rm = TRUE),
    max_remodeled = max(year_remodeled, na.rm = TRUE)
  )
```
We additionally check if the year remodeled is not 0 it is always more recent than
the year built.
```{r}
impossible_dates <- df_all %>%
  filter(year_remodeled > 0 & year_remodeled < year_built)

nrow(impossible_dates)
```
Check the valuation range.

```{r}
#| label: tbl-valuation-range
#| tbl-cap: "Property valuation range in current year market values (in USD)"

df_all %>%
  summarise(
    min_value     = min(total_market_value_current_year, na.rm = TRUE),
    max_value     = max(total_market_value_current_year, na.rm = TRUE)
  )
```

```{r}
#save(df_all, file = here::here("data", "df_all.RData"))
```
