---
title: "KNN model"
format:
  html:
    toc: true
    number-sections: true
---

# Introduction

This document provides a detailed interpretation of the KNN model to predict residential
property prices in Pierce County. The analysis builds upon the data prepared during the exploratory phase and focuses on identifying the key drivers of house prices and assessing the predictive power of the resulting models.

# Step-by-Step Analysis

```{r}
#| label: setup
#| message: false
#| warning: false

library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(tidyr)
library(here)

# Load the prepared data
model_df <- read_csv(here("..", "..", "data", "house_price_model_data.csv"))
```

```{r}
library(dotenv)
library(here)
library(caret)
library(tidyverse)
library(kknn)
options(scipen = 999)
```


## Feature selection
```{r}
load("../../df_all.RData")
str(df_all)
```
## Minimal model
We will first attempt to fit a KNN model with using only longitude, latitude and
the net square footage of properties.
```{r}
baseline_data <- df_all %>% 
  select(
    total_market_value_current_year, 
    latitude, 
    longitude, 
    land_net_squarefeet
  ) %>% 
  drop_na() # in case, but no NA

set.seed(222) 
train_index <- createDataPartition(
  baseline_data$total_market_value_current_year, 
  p = 0.8, 
  list = FALSE
)

train_data_1 <- baseline_data[train_index, ]
test_data_1  <- baseline_data[-train_index, ]
```

We try for different values of k (2, 3, 4, 5). We decided these values as in a map
visualization (showing the closest observations to the user) more than 5 similar properties displayed would compromise the user experience.
```{r}

cv_control <- trainControl(
  method = "cv", 
  number = 3
)

k_grid <- data.frame(k = c(2, 3, 4, 5))

set.seed(222)
knn_baseline <- train(
  total_market_value_current_year ~ latitude + longitude + land_net_squarefeet,
  data = train_data_1,
  method = "knn",                  # Back to the fast, standard method
  trControl = cv_control,
  preProcess = c("center", "scale"),
  tuneGrid = k_grid
)

print(knn_baseline)

```
The optimal number of k is 5. This model has an RMSE of 185 952 dollars on validation
set. The R-squared is almost 0.65, which considering that we only used the property
coordinates and the net square footage, is a great result.
```{r}
# Predict on the test data
predictions_1 <- predict(knn_baseline, newdata = test_data_1)

# Calculate performance metrics (RMSE, R-squared, MAE)
evaluation_metrics_1 <- postResample(pred = predictions_1, obs = test_data_1$total_market_value_current_year)
print(evaluation_metrics_1)

```
The MAE is 91 586 dollars on the teest set, a really good result for using only
2 features.

```{r}
train_preds_1 <- predict(knn_baseline, newdata = train_data_1)

# Absolute Percentage Error (APE) for training data
train_ape_1 <- abs(train_preds_1 - train_data_1$total_market_value_current_year) / train_data_1$total_market_value_current_year

train_within_15_1 <- mean(train_ape_1 <= 0.15) * 100

# Absolute Percentage Error (APE) for test data
test_ape_1 <- abs(predictions_1 - test_data_1$total_market_value_current_year) / test_data_1$total_market_value_current_year

test_within_15_1 <- mean(test_ape_1 <= 0.15) * 100

print(c(train_within_15_1, test_within_15_1))

```


On the minimal model almost 78% of the training data is predicted within 15% of
the actual price, while in the test data almost 71%. This is great considering that
we only used latitude, longitude and net square footage as features. However, our
objective is to have 80% within 15% of the actual value so we will try to get closer
by adding more features to the model.

## Adding features

```{r}
features_to_test <- df_all %>% 
  select(
    total_market_value_current_year,
    stories, bedrooms, bathrooms, units, year_built, year_remodeled
  ) %>% 
  drop_na() # in case, but here no NAs

cor_matrix <- cor(features_to_test)

price_correlations <- cor_matrix[, "total_market_value_current_year"] %>% 
  sort(decreasing = TRUE)

price_correlations
```
We will first try to only add one feature, the one with the highest correlation to
market value. This is the number of bathrooms.

```{r}
expanded_data <- df_all %>% 
  select(
    total_market_value_current_year, 
    latitude, 
    longitude, 
    land_net_squarefeet,
    bathrooms # newly added feature
  ) %>% 
  drop_na() # none, but just in case


train_data_2 <- expanded_data[train_index, ]
test_data_2  <- expanded_data[-train_index, ]
```

```{r}

set.seed(222)
knn_v2 <- train(
  total_market_value_current_year ~ latitude + longitude + land_net_squarefeet + bathrooms,
  data = train_data_2,
  method = "knn", 
  trControl = cv_control,
  preProcess = c("center", "scale"), 
  tuneGrid = k_grid
)

print(knn_v2)
```
Again the optimal value of k is 5. The RMSE is 186 373 a slight decrease from the
minimal model (185 952). The R-squared is about the same as for the minimal model.


```{r}
predictions_2 <- predict(knn_v2, newdata = test_data_2)

# Calculate performance metrics (RMSE, R-squared, MAE)
evaluation_metrics_2 <- postResample(pred = predictions_2, obs = test_data_2$total_market_value_current_year)
print(evaluation_metrics_2)
```
The MAE decreased from the previous model by almost 10 000, the RMSE by over 5 000.

```{r}
test_ape_2 <- abs(predictions_2 - test_data_2$total_market_value_current_year) / test_data_2$total_market_value_current_year
test_within_15_2 <- mean(test_ape_2 <= 0.15) * 100

train_preds_2 <- predict(knn_v2, newdata = train_data_2)
train_ape_2 <- abs(train_preds_2 - train_data_2$total_market_value_current_year) / train_data_2$total_market_value_current_year
train_within_15_2 <- mean(train_ape_2 <= 0.15) * 100

print(c(train_within_15_2, test_within_15_2))
```

To our custom benchmark, by adding bathrooms to this model, it performs better on
the train set, namely over 82% of the test data is predicted within 15% from the 
actual market value (previous model: 78%). Interestingly the performance on the 
training data is slightly worse, 78% compared to 78% previously.

### Another feature addition
We add another variable with the second highest correlation to the taregt variable.
This is the building year of the house.
```{r}
expanded_data_2 <- df_all %>% 
  select(
    total_market_value_current_year, 
    latitude, 
    longitude, 
    land_net_squarefeet,
    bathrooms,
    year_built  # newly added feature
  ) %>% 
  drop_na() # none, but just in case

train_data_3 <- expanded_data_2[train_index, ]
test_data_3  <- expanded_data_2[-train_index, ]
```

```{r}

set.seed(222)
knn_v3 <- train(
  total_market_value_current_year ~ latitude + longitude + land_net_squarefeet + bathrooms
  + year_built,
  data = train_data_3,
  method = "knn", 
  trControl = cv_control,
  preProcess = c("center", "scale"), 
  tuneGrid = k_grid
)

print(knn_v3)
```
Again the optimal value for k is 5. Both the RMSE and the R-square has worsened
compared to the previous model.

```{r}
predictions_3 <- predict(knn_v3, newdata = test_data_3)

# Calculate performance metrics (RMSE, R-squared, MAE)
evaluation_metrics_3 <- postResample(pred = predictions_3, obs = test_data_3$total_market_value_current_year)
print(evaluation_metrics_3)
```

Adding the year built yield worse results than the previous model that was expanded
with the number of bathrooms.

```{r}
test_ape_3 <- abs(predictions_3 - test_data_3$total_market_value_current_year) / test_data_3$total_market_value_current_year
test_within_15_3 <- mean(test_ape_3 <= 0.15) * 100

train_preds_3 <- predict(knn_v3, newdata = train_data_3)
train_ape_3 <- abs(train_preds_3 - train_data_3$total_market_value_current_year) / train_data_3$total_market_value_current_year
train_within_15_3 <- mean(train_ape_3 <= 0.15) * 100

print(c(train_within_15_3, test_within_15_3))
```
In this model only 81% of predictions in the test data are within 15% of the actual value (previous model: 82%). The performance on the training data also decreased to 75%
compared to 76% previously.

### All features

For the sake of seeing how much explanation on preciseness we are trading for better
user experience, we will fit a model that uses all numeric features that are expected
to be know by the user.

```{r}
df_full <- df_all %>% 
  select(
    total_market_value_current_year, 
    latitude, 
    longitude, 
    land_net_squarefeet,
    bedrooms,
    bathrooms,
    stories,
    units,
    year_built,
    year_remodeled
  ) %>% 
  drop_na() # none, but just in case

train_data_4 <- df_full[train_index, ]
test_data_4  <- df_full[-train_index, ]
```

```{r}
set.seed(222)
knn_full <- train(
  total_market_value_current_year ~ latitude + longitude + land_net_squarefeet + bathrooms
  + bedrooms + stories + units + year_built + year_remodeled,
  data = train_data_4,
  method = "knn", 
  trControl = cv_control,
  preProcess = c("center", "scale"), 
  tuneGrid = data.frame(k=5)
)

print(knn_full)
```
When all available numeric features were used, the RMSE was the highest, over
200 000 dollars. At the same time the R-squared dropped to 0.56.

```{r}
predictions_full <- predict(knn_full, newdata = test_data_4)

# Calculate performance metrics (RMSE, R-squared, MAE)
evaluation_metrics_full <- postResample(pred = predictions_full, obs = test_data_4$total_market_value_current_year)
print(evaluation_metrics_full)
```
The performance on the test data is the worst for the full model.

```{r}
test_ape_full <- abs(predictions_full - test_data_4$total_market_value_current_year) / test_data_4$total_market_value_current_year
test_within_15_full <- mean(test_ape_full <= 0.15) * 100

train_preds_full <- predict(knn_full, newdata = train_data_4)
train_ape_full <- abs(train_preds_full - train_data_4$total_market_value_current_year) / train_data_4$total_market_value_current_year
train_within_15_full <- mean(train_ape_full <= 0.15) * 100

print(c(train_within_15_full, test_within_15_full))
```
The large model performs quite well on the test data, predicting 80% of the observations
within 15% of the actual value. However, on the training set the performance is 
significantly worse (and the weakest among all models), predicting only 74% of the
observations within 15% of the the actual value.

## Model Comparison

| Model | RMSE | R-squared | MAE | within 15% test | within 15% train
|:----------|:-------|:-------|:-------|
| **minimal** | `r evaluation_metrics_1[["RMSE"]]` | `r evaluation_metrics_1[["Rsquared"]]` | `r evaluation_metrics_1[["MAE"]]` |
| **+bathrooms** | `r evaluation_metrics_2[["RMSE"]]`  | `r evaluation_metrics_2[["Rsquared"]]` | `r evaluation_metrics_2[["MAE"]]` |
| **+bathrooms+year_built** | `r evaluation_metrics_3[["RMSE"]]`  | `r evaluation_metrics_3[["Rsquared"]]` | `r evaluation_metrics_3[["MAE"]]` |
| **full** | `r evaluation_metrics_full[["RMSE"]]` | `r evaluation_metrics_full[["Rsquared"]]` | `r evaluation_metrics_full[["MAE"]]` |


| Model | within 15% train | within 15% test
|:----------|:-------|:-------|
| **minimal** | `r train_within_15_1` | `r test_within_15_1` |
| **+bathrooms** | `r train_within_15_2` | `r test_within_15_3` |
| **+bathrooms+year_built** | `r train_within_15_3` | `r test_within_15_3` |
| **full** | `r train_within_15_full` | `r test_within_15_full` |


