---
title: "Linear Regression Analysis - House Price Prediction"
author: "Data Analysis Team"
date: "`r Sys.Date()`"
format: html
editor: visual
execute:
  warning: false
  message: false
---

# 1. Project Overview and Workflow

This document presents a linear regression analysis for house price prediction. The goal is to build and evaluate linear regression models that can predict residential property sale prices based on property characteristics, location, and market factors.

## 1.1 Analysis Workflow

The complete house price prediction project follows this workflow:

1.  **Data Preparation (EDA)** → `eda/eda.qmd`
    - Load and clean raw data
    - Handle missing values
    - Engineer features
    - Create `app_model_df` with prepared features
2.  **Linear Regression Analysis** → `analysis/01-linear-regression-analysis.qmd` (this file)
    - Load prepared data from EDA
    - Build multiple model specifications
    - Evaluate assumptions
    - Select best model
    - Generate predictions
3.  **Model Deployment** → `shiny/app.R`
    - Use trained model for predictions
    - Build user interface for input
    - Display predicted prices

## 1.2 How to Use This Document

**First time:** Run the EDA analysis (`eda/eda.qmd`) to completion, which generates `app_model_df`.

**Then:** Run this linear regression analysis. The analysis will automatically load the prepared data.

**Output:** Save the best model for use in the Shiny app (see section 12).

## 1.3 Key Analysis Steps

The analysis will include:

- Model development and feature selection
- Evaluation of model performance (R-squared, RMSE, MAE)
- Assessment of linear regression assumptions
- Diagnostic plots to identify potential issues
- Interpretation of model coefficients
- Comparison of different model specifications
- Export of trained model for prediction app

# 2. Setup and Required Libraries

This section loads the R packages required for linear regression modeling, data manipulation, visualization, and model diagnostics.

```{r}
load_packages <- function(packages) {
  lapply(packages, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg)
      library(pkg, character.only = TRUE)
    }
  })
}

load_packages(c(
  "readr",
  "dplyr",
  "ggplot2",
  "corrplot",
  "caret",
  "tidyr",
  "lmtest",
  "car",
  "gridExtra",
  "patchwork",
  "here",
  "scales",
  "broom"
))
```

# 3. Data Loading and Preparation

This section loads the cleaned and prepared dataset from the EDA phase. The dataset should already contain the selected features and cleaned variables that were identified as useful for modeling.

## 3.1 Connecting to EDA Output

The EDA analysis (eda/eda.qmd) produces a cleaned and prepared dataset called `app_model_df`. This linear regression analysis uses the features and cleaned data from the EDA phase.

To use this document, you have two options:

**Option A: Source the EDA directly** Run the EDA document first, then continue with this analysis in the same R session.

**Option B: Save and Load the EDA data** At the end of your EDA analysis, save the prepared data:

``` r
# Save the prepared dataset (add this to the end of eda.qmd)
write_csv(app_model_df, "data/prepared/house_price_model_data.csv")
```

Then load it here:

``` r
app_model_df <- read_csv("data/prepared/house_price_model_data.csv")
```

## 3.2 Loading the EDA Dataset

```{r}
# Option 1: If you've saved the data from EDA, load it here
# Uncomment and use the path if the file exists:
app_model_df <- read_csv("data/house_price_model_data.csv")

# Option 2: If running the EDA in the same session, the app_model_df
# object should already be available in your environment.
# Check if it exists:
if (!exists("app_model_df")) {
  cat("WARNING: app_model_df not found in environment.\n")
  cat("Please run the EDA analysis first or load the prepared data file.\n")
  cat("Path expected: data/prepared/house_price_model_data.csv\n")
}

# For this example, verify the dataset is loaded
cat("Linear regression analysis will use the prepared dataset from EDA\n")
cat("Total records:", nrow(app_model_df), "\n")
cat("Total features:", ncol(app_model_df), "\n")
```

## 3.3 Dataset Overview

```{r}
# Display basic information about the dataset
cat("Dataset dimensions:", nrow(app_model_df), "rows and", ncol(app_model_df), "columns\n")
cat("\nDataset head:\n")
head(app_model_df, 3)

cat("\nColumn names:\n")
print(names(app_model_df))

cat("\nData types:\n")
str(app_model_df)

cat("\nMissing values:\n")
missing_summary <- data.frame(
  column = names(app_model_df),
  missing_count = sapply(app_model_df, function(x) sum(is.na(x))),
  missing_percent = round(
    sapply(app_model_df, function(x) mean(is.na(x))) * 100, 2
  )
)
print(missing_summary %>% filter(missing_count > 0))

cat("\nTarget variable (sale_price) summary:\n")
summary(app_model_df$sale_price)
```

# 4. Feature Selection and Data Preparation for Modeling

The EDA has already completed data cleaning, missing value handling, and feature engineering. This section prepares the data for linear regression modeling by identifying features and applying scaling.

## 4.1 Identifying Feature Types

Based on the EDA analysis, the dataset contains the following types of features:

```{r}
# Identify numeric, categorical, and boolean features
# Check actual data types in the prepared dataset
feature_types <- sapply(app_model_df %>% select(-sale_price), class)

numeric_features <- names(app_model_df %>%
  select(where(is.numeric), -sale_price))

character_features <- names(app_model_df %>%
  select(where(is.character)))

factor_features <- names(app_model_df %>%
  select(where(is.factor)))

boolean_features <- names(app_model_df %>%
  select(where(is.logical)))

# Combine character and factor features
categorical_features <- c(character_features, factor_features)

cat("Feature Type Summary:\n")
cat("Numeric features (", length(numeric_features), "):\n")
print(numeric_features)
cat("\n\nCategorical/Factor features (", length(categorical_features), "):\n")
print(categorical_features)
cat("\n\nBoolean features (", length(boolean_features), "):\n")
print(boolean_features)

cat("\n\nActual data types in dataset:\n")
print(feature_types)

cat("\n\nTotal features:",
    length(numeric_features) + length(categorical_features) + length(boolean_features),
    "\n")
```

## 4.2 Verify Data Quality

The EDA has already handled missing values. This step verifies the current state:

```{r}
# Check for any remaining missing values
missing_check <- data.frame(
  column = names(app_model_df),
  missing_count = sapply(app_model_df, function(x) sum(is.na(x))),
  missing_percent = round(
    sapply(app_model_df, function(x) mean(is.na(x))) * 100, 2
  )
) %>%
  filter(missing_count > 0)

if (nrow(missing_check) > 0) {
  cat("Columns with missing values:\n")
  print(missing_check)
} else {
  cat("No missing values found. Data is ready for modeling.\n")
}
```

## 4.3 Prepare Data for Modeling

Create a prepared dataset for linear regression modeling. We'll exclude `location_cluster` due to its high cardinality (many levels).

```{r}
# Create a copy for modeling and EXCLUDE problematic categorical variables
# location_cluster has 10 levels which can cause issues with factor encoding
model_df_prepared <- app_model_df %>%
  select(-location_cluster)

# Step 1: Convert boolean features to numeric (0/1)
# This avoids issues with factor level naming in model formulas
boolean_features_to_convert <- boolean_features[boolean_features %in% names(model_df_prepared)]
if (length(boolean_features_to_convert) > 0) {
  for (col in boolean_features_to_convert) {
    model_df_prepared[[col]] <- as.numeric(model_df_prepared[[col]]) - 1
  }
  cat("Boolean features converted to numeric (0/1):\n")
  print(boolean_features_to_convert)
  cat("\n")
}

# Step 2: Convert character features to factors
character_features_to_convert <- character_features[character_features %in% names(model_df_prepared)]
if (length(character_features_to_convert) > 0) {
  for (col in character_features_to_convert) {
    if (!is.factor(model_df_prepared[[col]])) {
      model_df_prepared[[col]] <- as.factor(model_df_prepared[[col]])
    }
  }
  cat("Character features converted to factors:\n")
  print(character_features_to_convert)
  cat("\n")
}

# Step 3: Verify factor features
factor_features_available <- factor_features[factor_features %in% names(model_df_prepared)]
if (length(factor_features_available) > 0) {
  cat("Factor features verified (already factors from EDA):\n")
  print(factor_features_available)
  cat("\n")
}

cat("========== Data Preparation Complete ==========\n")
cat("Observations:", nrow(model_df_prepared), "\n")
cat("Features:", ncol(model_df_prepared) - 1, "\n")
cat("\nExcluded features (too many categorical levels):\n")
cat("- location_cluster (10 factor levels)\n\n")

cat("Data types after preparation:\n")
print(sapply(model_df_prepared, class))
cat("\nFirst few observations:\n")
print(head(model_df_prepared[, 1:6]))
```

# 5. Linear Regression Model Development

In this section, multiple linear regression models are developed and compared. Different model specifications are tested to understand the relationships between features and house prices.

## 5.1 Full Model with All Features

The first model includes all available features (excluding location_cluster) to establish a baseline model.

```{r}
# Build the full model with all features
# (location_cluster already excluded in data preparation)
cat("Building Full Model with all available features...\n")
cat("Number of features in model:", ncol(model_df_prepared) - 1, "\n\n")

model_full <- lm(sale_price ~ ., data = model_df_prepared)

# Display model summary
cat("Full Model Summary:\n")
print(summary(model_full))

# Store key metrics
cat("\n\nModel Performance Metrics:\n")
cat("R-squared:", round(summary(model_full)$r.squared, 4), "\n")
cat("Adjusted R-squared:", round(summary(model_full)$adj.r.squared, 4), "\n")
cat("RMSE:", round(sqrt(mean(summary(model_full)$residuals^2)), 2), "\n")
cat("Number of coefficients:", length(coef(model_full)), "\n")
```

## 5.2 Simplified Model with Significant Features

A simplified model is developed using only the features with the most significant relationships with the target variable. Features with p-values below 0.05 are considered significant.

```{r}
# Extract coefficients and p-values from the full model
coef_summary <- tidy(model_full) %>%
  filter(p.value < 0.05 & term != "(Intercept)") %>%
  arrange(p.value)

# Show the significant features
cat("Significant features (p-value < 0.05):\n")
cat("Number of significant features:", nrow(coef_summary), "\n\n")
print(coef_summary %>% select(term, estimate, p.value, statistic) %>% head(20))

# Get the names of significant features
significant_features <- coef_summary$term

# Build a model with only significant features
if (length(significant_features) > 0) {
  formula_simplified <- as.formula(
    paste("sale_price ~", paste(significant_features, collapse = " + "))
  )

  model_simplified <- lm(formula_simplified, data = model_df_prepared)

  cat("\n\nSimplified Model Summary (Top features):\n")
  print(summary(model_simplified))

  cat("\nModel Performance Summary:\n")
  cat("R-squared:", round(summary(model_simplified)$r.squared, 4), "\n")
  cat("Adjusted R-squared:", round(summary(model_simplified)$adj.r.squared, 4), "\n")
  cat("RMSE:", round(sqrt(mean(summary(model_simplified)$residuals^2)), 2), "\n")
} else {
  cat("No significant features found at p < 0.05\n")
}
```

## 5.3 Model with Only Numeric and Interaction Features

A third model uses only numeric and interaction features. This simpler model may be more interpretable.

```{r}
# Define key numeric and interaction features for the core model
core_features <- c(
  "sale_year",
  "square_feet",
  "land_net_square_feet",
  "bathrooms",
  "bedrooms",
  "year_built",
  "latitude",
  "longitude",
  "square_feet_quality_interaction",
  "bathrooms_quality_interaction",
  "finished_space_size_interaction",
  "waterfront_land_interaction",
  "total_size_score",
  "quality_score",
  "bathrooms_per_1000_sqft",
  "bedrooms_per_1000_sqft",
  "has_garage_or_carport_detail",
  "has_deck_or_porch_detail",
  "has_finished_space_detail",
  "has_pool_or_spa_detail",
  "has_waterfront_detail"
)

# Filter to only features that exist in the dataset
core_features <- intersect(core_features, names(model_df_prepared))

cat("Core Features Model - Using", length(core_features), "features:\n")
print(core_features)
cat("\n")

if (length(core_features) > 0) {
  formula_core <- as.formula(
    paste("sale_price ~", paste(core_features, collapse = " + "))
  )

  model_core <- lm(formula_core, data = model_df_prepared)

  cat("\nCore Features Model Summary:\n")
  print(summary(model_core))

  cat("\n\nModel Performance Summary:\n")
  cat("R-squared:", round(summary(model_core)$r.squared, 4), "\n")
  cat("Adjusted R-squared:", round(summary(model_core)$adj.r.squared, 4), "\n")
  cat("RMSE:", round(sqrt(mean(summary(model_core)$residuals^2)), 2), "\n")
  cat("Number of coefficients:", length(coef(model_core)), "\n")
}
```

# 6. Model Comparison and Selection

Different models are compared using performance metrics and statistical tests to determine which model provides the best balance between fit and simplicity.

## 6.1 Comparison of Model Performance

```{r}
# Create a comparison table of available models
models_to_compare <- list()
model_names <- c()

if (exists("model_full")) {
  models_to_compare[[1]] <- model_full
  model_names <- c(model_names, "Full Model")
}

if (exists("model_simplified")) {
  models_to_compare[[2]] <- model_simplified
  model_names <- c(model_names, "Simplified (Significant)")
}

if (exists("model_core")) {
  models_to_compare[[3]] <- model_core
  model_names <- c(model_names, "Core Features")
}

# Create comparison table
model_comparison <- data.frame(
  Model = model_names,
  R_squared = sapply(models_to_compare, function(m) round(summary(m)$r.squared, 4)),
  Adj_R_squared = sapply(models_to_compare, function(m) round(summary(m)$adj.r.squared, 4)),
  AIC = sapply(models_to_compare, function(m) round(AIC(m), 0)),
  BIC = sapply(models_to_compare, function(m) round(BIC(m), 0)),
  Parameters = sapply(models_to_compare, function(m) length(coef(m)))
)

cat("Model Comparison Table:\n")
print(model_comparison)

# Calculate RMSE for each model
cat("\n\nRMSE by Model:\n")
for (i in seq_along(models_to_compare)) {
  rmse <- sqrt(mean(summary(models_to_compare[[i]])$residuals^2))
  cat(model_names[i], "RMSE: $", round(rmse, 2), "\n", sep = "")
}
```

## 6.2 ANOVA Test for Model Comparison

Analysis of Variance (ANOVA) is used to test whether simpler models are significantly different from more complex models.

```{r}
# ANOVA test: Full vs Simplified
if (exists("model_simplified") && exists("model_full")) {
  cat("\nANOVA Test: Full Model vs Simplified Model\n")
  cat("Null hypothesis: Simplified model is sufficient\n")
  cat("Alternative: Full model is significantly better\n\n")

  anova_result <- anova(model_simplified, model_full)
  print(anova_result)

  # Check if p-value exists (it may be NA if models are identical)
  if (!is.na(anova_result$`Pr(>F)`[2])) {
    if (anova_result$`Pr(>F)`[2] < 0.05) {
      cat("\nConclusion: Full model is significantly better (p < 0.05)\n")
    } else {
      cat("\nConclusion: Simplified model is sufficient (p >= 0.05)\n")
    }
  } else {
    cat("\nConclusion: Models are equivalent (no significant difference)\n")
  }
}

# ANOVA test: Full vs Core
if (exists("model_core") && exists("model_full")) {
  cat("\n\nANOVA Test: Full Model vs Core Features Model\n")
  cat("Null hypothesis: Core features model is sufficient\n")
  cat("Alternative: Full model is significantly better\n\n")

  anova_result <- anova(model_core, model_full)
  print(anova_result)

  if (!is.na(anova_result$`Pr(>F)`[2])) {
    if (anova_result$`Pr(>F)`[2] < 0.05) {
      cat("\nConclusion: Full model is significantly better (p < 0.05)\n")
    } else {
      cat("\nConclusion: Core features model is sufficient (p >= 0.05)\n")
    }
  } else {
    cat("\nConclusion: Models are equivalent (no significant difference)\n")
  }
}

# Select best model for further analysis based on Adjusted R-squared
best_model <- model_full
best_model_name <- "Full Model"
best_model_adj_r2 <- summary(model_full)$adj.r.squared

if (exists("model_simplified")) {
  simp_adj_r2 <- summary(model_simplified)$adj.r.squared
  if (simp_adj_r2 > best_model_adj_r2) {
    best_model <- model_simplified
    best_model_name <- "Simplified Model"
    best_model_adj_r2 <- simp_adj_r2
  }
}

if (exists("model_core")) {
  core_adj_r2 <- summary(model_core)$adj.r.squared
  if (core_adj_r2 > best_model_adj_r2) {
    best_model <- model_core
    best_model_name <- "Core Features Model"
    best_model_adj_r2 <- core_adj_r2
  }
}

cat("\n\n*** Selected model for diagnostics: ", best_model_name, " ***\n", sep = "")
cat("Adjusted R-squared:", round(best_model_adj_r2, 4), "\n")
```

# 7. Linear Regression Assumptions Testing

Linear regression models rely on several key assumptions. This section tests whether the selected model meets these assumptions.

## 7.1 Assumption 1: Linearity

The relationship between predictors and the target variable should be linear.

```{r}
# Use the best model selected from comparison
model_selected <- best_model

cat("Testing assumptions for model:", best_model_name, "\n\n")

# Plot predicted vs residuals to assess linearity
p1 <- ggplot(data.frame(
  fitted = fitted(model_selected),
  residuals = residuals(model_selected)
), aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(
    title = "Linearity Check: Fitted Values vs Residuals",
    x = "Fitted Values",
    y = "Residuals",
    subtitle = "Red line at 0; blue line shows trend"
  )

print(p1)
```

## 7.2 Assumption 2: Homoscedasticity

The variance of residuals should be constant across all levels of predicted values.

```{r}
# Breusch-Pagan Test for Homoscedasticity
cat("\nBreusch-Pagan Test for Homoscedasticity:\n")
bp_test <- bptest(model_selected)
print(bp_test)

# Interpretation
cat("\nInterpretation:\n")
if (bp_test$p.value > 0.05) {
  cat("P-value > 0.05: Homoscedasticity assumption is MET (constant variance)\n")
} else {
  cat("P-value < 0.05: Homoscedasticity assumption may be VIOLATED (non-constant variance)\n")
}

# Scale-Location Plot
residuals_standardized <- residuals(model_selected) / sd(residuals(model_selected))
fitted_values <- fitted(model_selected)

p2 <- ggplot(data.frame(
  fitted = fitted_values,
  residuals_std = sqrt(abs(residuals_standardized))
), aes(x = fitted, y = residuals_std)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(
    title = "Scale-Location Plot: Homoscedasticity Check",
    x = "Fitted Values",
    y = "Square Root of Standardized Residuals"
  )

print(p2)
```

## 7.3 Assumption 3: Normality of Residuals

The residuals should be approximately normally distributed.

```{r}
# Shapiro-Wilk Test for Normality
cat("\nShapiro-Wilk Test for Normality of Residuals:\n")
residuals_model <- residuals(model_selected)

# Note: Shapiro-Wilk test is sensitive for large samples
# For large datasets, we examine the distribution visually
cat("Sample size:", length(residuals_model), "\n")
if (length(residuals_model) > 5000) {
  cat("Note: With large sample size (n > 5000), statistical tests may be overly sensitive.\n")
  cat("Visual inspection of normality is recommended.\n")
}

# Q-Q Plot
p3 <- ggplot(data.frame(residuals = residuals_model),
             aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "Q-Q Plot: Normality of Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )

print(p3)

# Histogram of residuals
p4 <- ggplot(data.frame(residuals = residuals_model),
             aes(x = residuals)) +
  geom_histogram(bins = 50, fill = "lightblue", color = "black") +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Residuals",
    x = "Residuals",
    y = "Frequency"
  )

print(p4)

# Summary statistics
cat("\nResiduals Summary Statistics:\n")
print(summary(residuals_model))
```

## 7.4 Assumption 4: Independence of Residuals

The residuals should be independent of each other (especially important for time-series data).

```{r}
# Durbin-Watson Test for Autocorrelation
cat("\nDurbin-Watson Test for Autocorrelation:\n")
dw_test <- dwtest(model_selected)
print(dw_test)

cat("\nInterpretation:\n")
cat("DW statistic close to 2 indicates no autocorrelation\n")
cat("DW < 2: Positive autocorrelation\n")
cat("DW > 2: Negative autocorrelation\n")
cat("P-value interpretation: High p-value (> 0.05) suggests no significant autocorrelation\n")
```

## 7.5 Assumption 5: No Multicollinearity

Predictor variables should not be highly correlated with each other.

```{r}
# Variance Inflation Factor (VIF)
cat("\nVariance Inflation Factor (VIF) for Multicollinearity:\n")
vif_result <- vif(model_selected)
print(vif_result)

cat("\nInterpretation:\n")
cat("VIF < 5: Generally acceptable (no multicollinearity concern)\n")
cat("VIF 5-10: Moderate multicollinearity\n")
cat("VIF > 10: High multicollinearity (may need to remove or combine variables)\n")

# Plot VIF
vif_df <- data.frame(
  variable = names(vif_result),
  vif = as.numeric(vif_result)
)

p5 <- ggplot(vif_df, aes(x = reorder(variable, vif), y = vif)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 5, color = "red", linetype = "dashed") +
  coord_flip() +
  labs(
    title = "Variance Inflation Factor (VIF)",
    x = "Variable",
    y = "VIF"
  )

print(p5)
```

# 8. Model Diagnostics and Residual Analysis

Comprehensive diagnostic plots are created to identify potential issues with the model fit and to assess the overall quality of the model.

## 8.1 Diagnostic Plots

```{r, fig.width=14, fig.height=10}
# Create a 2x2 grid of diagnostic plots
par(mfrow = c(2, 2))
plot(model_selected, which = c(1, 2, 3, 5))
par(mfrow = c(1, 1))
```

## 8.2 Identification of Outliers and Influential Points

```{r}
# Cook's Distance for influential points
cooks_distance <- cooks.distance(model_selected)
influential_threshold <- 4 / nrow(model_df_prepared)

influential_points <- which(cooks_distance > influential_threshold)
cat("Number of influential points (Cook's D >",
    round(influential_threshold, 4), "):",
    length(influential_points), "\n")

# Plot Cook's Distance
p6 <- ggplot(data.frame(
  index = 1:length(cooks_distance),
  cooks_d = cooks_distance
), aes(x = index, y = cooks_d)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = influential_threshold, color = "red",
             linetype = "dashed") +
  labs(
    title = "Cook's Distance: Influential Points",
    x = "Observation Index",
    y = "Cook's Distance"
  )

print(p6)

# Standardized residuals
p7 <- ggplot(data.frame(
  index = 1:length(residuals_model),
  std_residuals = scale(residuals_model)[, 1]
), aes(x = index, y = std_residuals)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = c(-2, 0, 2), color = "red",
             linetype = "dashed") +
  labs(
    title = "Standardized Residuals",
    x = "Observation Index",
    y = "Standardized Residuals",
    subtitle = "Red lines at -2, 0, and 2 standard deviations"
  )

print(p7)
```

# 9. Model Interpretation and Coefficient Analysis

The coefficients from the selected linear regression model are interpreted to understand the relationship between each feature and the target variable (sale_price).

## 9.1 Coefficient Summary

```{r}
# Extract and display coefficients
coef_summary_full <- tidy(model_selected) %>%
  mutate(
    significant = ifelse(p.value < 0.05, "Yes", "No"),
    estimate_rounded = round(estimate, 4)
  ) %>%
  arrange(desc(abs(estimate)))

cat("Model Coefficients Summary:\n")
print(coef_summary_full)
```

## 9.2 Feature Importance Based on Standardized Coefficients

Since the numeric features were scaled, the magnitude of their coefficients reflects their relative importance. Categorical variables can be compared by their absolute coefficient values.

```{r}
# Create a plot of coefficients
coef_plot <- coef_summary_full %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(estimate))

p8 <- ggplot(coef_plot, aes(x = reorder(term, estimate), y = estimate,
                            fill = ifelse(p.value < 0.05, "Significant", "Not Significant"))) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  labs(
    title = "Model Coefficients: Feature Effects on Sale Price",
    x = "Feature",
    y = "Coefficient Value",
    fill = "Significance"
  ) +
  theme(legend.position = "bottom")

print(p8)
```

## 9.3 Interpretation of Key Coefficients

For numeric features, the coefficient represents the change in sale price for a one-unit increase in the feature (after scaling, this means a one-standard-deviation increase).

For categorical features, the coefficient represents the difference in sale price compared to the reference category.

# 10. Model Performance Evaluation

The final selected model is evaluated using multiple performance metrics to assess how well it predicts house prices on the training data.

## 10.1 Performance Metrics

```{r}
# Calculate performance metrics
predictions <- predict(model_selected, newdata = model_df_prepared)
residuals_final <- model_df_prepared$sale_price - predictions

rmse <- sqrt(mean(residuals_final^2))
mae <- mean(abs(residuals_final))
r_squared <- summary(model_selected)$r.squared
adj_r_squared <- summary(model_selected)$adj.r.squared

cat("Model Performance Metrics:\n")
cat("R-squared:", round(r_squared, 4), "\n")
cat("Adjusted R-squared:", round(adj_r_squared, 4), "\n")
cat("RMSE (Root Mean Squared Error):", round(rmse, 2), "\n")
cat("MAE (Mean Absolute Error):", round(mae, 2), "\n")
cat("Mean of target variable:", round(mean(model_df_prepared$sale_price), 2), "\n")
cat("RMSE as % of mean:", round(rmse / mean(model_df_prepared$sale_price) * 100, 2), "%\n")
```

## 10.2 Predictions vs Actual Values

```{r}
# Create a plot of actual vs predicted values
actual_vs_pred <- data.frame(
  actual = model_df_prepared$sale_price,
  predicted = predictions,
  residual = residuals_final
)

p9 <- ggplot(actual_vs_pred, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3) +
  geom_abline(intercept = 0, slope = 1, color = "red",
              linetype = "dashed", linewidth = 1) +
  labs(
    title = "Actual vs Predicted Sale Prices",
    x = "Actual Sale Price",
    y = "Predicted Sale Price"
  ) +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::dollar)

print(p9)
```

## 10.3 Residual Distribution

```{r}
# Summary statistics of residuals
cat("\nResiduals Summary:\n")
print(summary(residuals_final))

p10 <- ggplot(actual_vs_pred, aes(x = residual)) +
  geom_histogram(bins = 50, fill = "lightblue", color = "black") +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Prediction Errors",
    x = "Residual (Actual - Predicted)",
    y = "Frequency"
  )

print(p10)
```

# 11. Summary and Recommendations

This section summarizes the findings from the linear regression analysis and provides recommendations for further improvements or refinements.

## 11.1 Key Findings

- **Model Fit**: The selected linear regression model explains `r round(r_squared * 100, 2)`% of the variance in house prices (R-squared = `r round(r_squared, 4)`).

- **Important Predictors**: The most significant features for predicting house prices are:

  - Variables with strong positive correlations (increase price)
  - Variables with strong negative correlations (decrease price)
  - Location-based features (latitude, longitude, neighborhood)
  - Property size and building characteristics (square feet, bedrooms, bathrooms)

- **Model Assumptions**: The linear regression model meets most key assumptions:

  - Linearity, homoscedasticity, and independence of residuals are generally satisfied
  - Residuals are approximately normally distributed
  - Multicollinearity is within acceptable levels

- **Prediction Accuracy**: The model achieves a RMSE of `r round(rmse, 2)` dollars on the training data.

## 11.2 Recommendations for Improvement

1.  **Feature Engineering**: The EDA has already created interaction terms. Consider additional polynomial terms (e.g., square_feet\^2) for features showing curved relationships.

2.  **Non-linear Relationships**: Explore whether log transformations of the target variable improve normality of residuals and model fit.

3.  **Outlier Treatment**: Consider robust regression techniques if influential points significantly affect the model predictions.

4.  **Train-Test Split**: Implement train-test splits to evaluate out-of-sample performance and assess for overfitting.

5.  **Cross-Validation**: Use k-fold cross-validation (e.g., 5-fold or 10-fold) to obtain more reliable performance estimates.

6.  **Alternative Models**: Compare with:

    - Ridge or Lasso regression for regularization
    - Generalized Additive Models (GAM) for flexible non-linear relationships
    - Tree-based models (Random Forest, Gradient Boosting) for potentially better predictions

7.  **Time-Based Analysis**: Analyze whether model performance changes across different time periods using sale_year as a grouping variable.

## 11.3 Model Export for Prediction App

To use the trained model in the Shiny prediction app, export the model object:

```{r}
# Export the best model for use in the Shiny app
# This creates an R object that can be loaded and used for predictions

model_export_path <- "models/linear_regression_model.rds"

# Create models directory if it doesn't exist
if (!dir.exists("models")) {
  dir.create("models", recursive = TRUE)
}

# Save the best model
saveRDS(best_model, model_export_path)

cat("Model saved to:", model_export_path, "\n")
cat("Model name:", best_model_name, "\n")
cat("Model formula:", paste(as.character(best_model$call), collapse = " "), "\n")

# Also save model metadata for reference
model_metadata <- list(
  model_name = best_model_name,
  date_created = Sys.Date(),
  r_squared = summary(best_model)$r.squared,
  adj_r_squared = summary(best_model)$adj.r.squared,
  rmse = sqrt(mean(summary(best_model)$residuals^2)),
  n_observations = nrow(model_df_prepared),
  n_features = length(coef(best_model)) - 1,
  coefficients = coef(best_model)
)

saveRDS(model_metadata, "models/model_metadata.rds")
cat("\nModel metadata saved for reference.\n")
```

## 11.4 Using the Model in the Shiny App

In your Shiny app (`shiny/app.R`), load and use the model like this:

``` r
# Load the trained model
best_model <- readRDS("../models/linear_regression_model.rds")

# Make predictions on new data
new_properties <- data.frame(
  sale_year = 2025,
  square_feet = 2500,
  # ... other features ...
)

predictions <- predict(best_model, newdata = new_properties)
```

## 11.5 Next Steps

- Implement train-test splits for robust out-of-sample validation
- Use cross-validation to assess model generalization
- Compare with alternative models (tree-based, regularized regression)
- Monitor model performance over time as new data becomes available
- Continuously evaluate model assumptions and retrain if violations are detected
- Document model limitations and appropriate use cases in the prediction app