---
title: "Advanced Models Comparison - House Price Prediction"
author: "Data Analysis Team"
date: "`r Sys.Date()`"
format: html
editor: visual
execute:
  warning: false
  message: false
---

# Advanced Model Improvements and Comparison

This document implements advanced techniques to improve upon the baseline linear regression model, including:

1. Log-transformation of target variable
2. Polynomial feature engineering
3. 10-fold cross-validation evaluation
4. Regularization methods (Ridge/Lasso)
5. Tree-based models (Random Forest, Gradient Boosting)

# Setup and Libraries

```{r}
load_packages <- function(packages) {
  lapply(packages, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, repos = "https://cloud.r-project.org")
      library(pkg, character.only = TRUE)
    }
  })
}

load_packages(c(
  "readr",
  "dplyr",
  "ggplot2",
  "caret",
  "glmnet",
  "randomForest",
  "gbm",
  "tidyr",
  "broom",
  "scales",
  "patchwork",
  "gridExtra",
  "stringr"
))
```

# Load Data

```{r}
app_model_df <- read_csv("data/house_price_model_data.csv")

cat("Dataset loaded successfully\n")
cat("Observations:", nrow(app_model_df), "\n")
cat("Features:", ncol(app_model_df), "\n")
```

# Prepare Data for Advanced Modeling

```{r}
model_df_prepared <- app_model_df %>%
  select(-location_cluster)

boolean_features <- names(model_df_prepared %>%
  select(where(is.logical)))

if (length(boolean_features) > 0) {
  for (col in boolean_features) {
    model_df_prepared[[col]] <- as.numeric(model_df_prepared[[col]]) - 1
  }
}

cat("Data prepared. Total features:", ncol(model_df_prepared) - 1, "\n")
```

# 1. Log-Transformation of Target Variable

```{r}
cat("===== 1. LOG-TRANSFORMATION ANALYSIS =====\n\n")

model_df_log <- model_df_prepared %>%
  mutate(log_sale_price = log(sale_price))

cat("Original sale_price distribution:\n")
print(summary(model_df_log$sale_price))

cat("\n\nLog-transformed sale_price distribution:\n")
print(summary(model_df_log$log_sale_price))

# Create visualization comparing distributions
p1 <- ggplot(model_df_log, aes(x = sale_price)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  labs(title = "Original Sale Price Distribution",
       x = "Sale Price ($)",
       y = "Frequency") +
  scale_x_continuous(labels = scales::dollar)

p2 <- ggplot(model_df_log, aes(x = log_sale_price)) +
  geom_histogram(bins = 50, fill = "darkgreen", color = "black") +
  labs(title = "Log-Transformed Sale Price Distribution",
       x = "Log(Sale Price)",
       y = "Frequency")

print(p1 / p2)
```

## Build Linear Regression with Log-Transformed Price

```{r}
model_log_lr <- lm(log_sale_price ~ . -sale_price, data = model_df_log)

cat("\nLog-Transformed Linear Regression Model Summary:\n")
print(summary(model_log_lr))

cat("\n\nLog-Model Performance Metrics:\n")
cat("R-squared:", round(summary(model_log_lr)$r.squared, 4), "\n")
cat("Adjusted R-squared:", round(summary(model_log_lr)$adj.r.squared, 4), "\n")
cat("RMSE (log scale):", round(sqrt(mean(summary(model_log_lr)$residuals^2)), 4), "\n")

# Convert back to original scale for interpretation
predictions_log <- predict(model_log_lr, newdata = model_df_log)
predictions_original <- exp(predictions_log)
actual <- model_df_log$sale_price

rmse_log_transformed <- sqrt(mean((actual - predictions_original)^2))
cat("RMSE (original scale): $", round(rmse_log_transformed, 2), "\n", sep = "")
cat("Improvement vs baseline ($187,104.50):",
    round(187104.50 - rmse_log_transformed, 2), "\n\n")
```

## Q-Q Plot for Log-Transformed Residuals

```{r}
residuals_log <- residuals(model_log_lr)

p_qq <- ggplot(data.frame(residuals = residuals_log),
               aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Log-Transformed Model Residuals",
       subtitle = "Improved normality compared to original scale",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles")

print(p_qq)
```

# 2. Polynomial Feature Engineering

```{r}
cat("\n===== 2. POLYNOMIAL FEATURE ENGINEERING =====\n\n")

model_df_poly <- model_df_prepared %>%
  mutate(
    square_feet_sq = square_feet^2,
    land_net_square_feet_sq = land_net_square_feet^2,
    bathrooms_sq = bathrooms^2,
    bedrooms_sq = bedrooms^2,
    year_built_sq = year_built^2,
    quality_score_sq = quality_score^2
  )

cat("Polynomial features created:\n")
cat("- square_feet_sq\n")
cat("- land_net_square_feet_sq\n")
cat("- bathrooms_sq\n")
cat("- bedrooms_sq\n")
cat("- year_built_sq\n")
cat("- quality_score_sq\n\n")

cat("New feature count:", ncol(model_df_poly) - 1, "\n")

# Build polynomial model
model_poly <- lm(sale_price ~ ., data = model_df_poly)

cat("\nPolynomial Model Summary:\n")
print(summary(model_poly))

cat("\n\nPolynomial Model Performance:\n")
cat("R-squared:", round(summary(model_poly)$r.squared, 4), "\n")
cat("Adjusted R-squared:", round(summary(model_poly)$adj.r.squared, 4), "\n")

predictions_poly <- predict(model_poly, newdata = model_df_poly)
residuals_poly <- model_df_poly$sale_price - predictions_poly
rmse_poly <- sqrt(mean(residuals_poly^2))
cat("RMSE: $", round(rmse_poly, 2), "\n", sep = "")
cat("Improvement vs baseline:", round(187104.50 - rmse_poly, 2), "\n\n")
```

## Comparison of Polynomial Terms Significance

```{r}
poly_coefs <- tidy(model_poly) %>%
  filter(str_detect(term, "_sq$")) %>%
  arrange(p.value)

cat("Polynomial Terms Significance (ordered by p-value):\n")
print(poly_coefs %>% select(term, estimate, p.value, statistic))
```

# 3. 10-Fold Cross-Validation

```{r}
cat("\n===== 3. 10-FOLD CROSS-VALIDATION =====\n\n")

set.seed(123)

# Define cross-validation setup
ctrl <- trainControl(
  method = "cv",
  number = 10,
  verboseIter = FALSE
)

# Model 1: Baseline Linear Regression
cat("Training Baseline Linear Regression with 10-fold CV...\n")
model_baseline_cv <- train(
  sale_price ~ .,
  data = model_df_prepared,
  method = "lm",
  trControl = ctrl
)

cat("\nBaseline Linear Regression - 10-Fold CV Results:\n")
print(model_baseline_cv)

# Model 2: Polynomial Linear Regression
cat("\n\nTraining Polynomial Linear Regression with 10-fold CV...\n")
model_poly_cv <- train(
  sale_price ~ .,
  data = model_df_poly,
  method = "lm",
  trControl = ctrl,
  preProcess = c("center", "scale")
)

cat("\nPolynomial Linear Regression - 10-Fold CV Results:\n")
print(model_poly_cv)
```

## Cross-Validation Results Comparison

```{r}
cv_results <- data.frame(
  Model = c("Baseline LR", "Polynomial LR"),
  RMSE_Mean = c(
    model_baseline_cv$results$RMSE[1],
    model_poly_cv$results$RMSE[1]
  ),
  RMSE_SD = c(
    model_baseline_cv$results$RMSESD[1],
    model_poly_cv$results$RMSESD[1]
  ),
  Rsquared_Mean = c(
    model_baseline_cv$results$Rsquared[1],
    model_poly_cv$results$Rsquared[1]
  )
)

cat("\n10-Fold Cross-Validation Summary:\n")
print(cv_results)

# Visualize CV results
cv_comparison <- data.frame(
  Model = rep(c("Baseline", "Polynomial"), each = 10),
  RMSE = c(
    model_baseline_cv$resample$RMSE,
    model_poly_cv$resample$RMSE
  )
)

p_cv <- ggplot(cv_comparison, aes(x = Model, y = RMSE, fill = Model)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "10-Fold Cross-Validation RMSE Comparison",
       y = "RMSE ($)") +
  theme(legend.position = "bottom")

print(p_cv)
```

# 4. Regularization Methods (Ridge and Lasso)

```{r}
cat("\n===== 4. REGULARIZATION METHODS: RIDGE & LASSO =====\n\n")

# Prepare data for glmnet (matrix format)
X <- model_df_prepared %>%
  select(-sale_price) %>%
  as.matrix()

y <- model_df_prepared$sale_price

# Ridge Regression
cat("Training Ridge Regression with CV...\n")
ridge_cv <- cv.glmnet(
  X, y,
  alpha = 0,
  nfolds = 10,
  type.measure = "mse"
)

cat("Ridge Regression - Best Lambda:", ridge_cv$lambda.min, "\n")

predictions_ridge <- predict(ridge_cv, newx = X, s = "lambda.min")[, 1]
rmse_ridge <- sqrt(mean((y - predictions_ridge)^2))
r2_ridge <- 1 - (sum((y - predictions_ridge)^2) / sum((y - mean(y))^2))

cat("Ridge RMSE: $", round(rmse_ridge, 2), "\n", sep = "")
cat("Ridge R-squared:", round(r2_ridge, 4), "\n\n")

# Lasso Regression
cat("Training Lasso Regression with CV...\n")
lasso_cv <- cv.glmnet(
  X, y,
  alpha = 1,
  nfolds = 10,
  type.measure = "mse"
)

cat("Lasso Regression - Best Lambda:", lasso_cv$lambda.min, "\n")

predictions_lasso <- predict(lasso_cv, newx = X, s = "lambda.min")[, 1]
rmse_lasso <- sqrt(mean((y - predictions_lasso)^2))
r2_lasso <- 1 - (sum((y - predictions_lasso)^2) / sum((y - mean(y))^2))

cat("Lasso RMSE: $", round(rmse_lasso, 2), "\n", sep = "")
cat("Lasso R-squared:", round(r2_lasso, 4), "\n\n")

# Elastic Net (combination)
cat("Training Elastic Net (alpha=0.5) with CV...\n")
elasticnet_cv <- cv.glmnet(
  X, y,
  alpha = 0.5,
  nfolds = 10,
  type.measure = "mse"
)

cat("Elastic Net - Best Lambda:", elasticnet_cv$lambda.min, "\n")

predictions_elasticnet <- predict(elasticnet_cv, newx = X, s = "lambda.min")[, 1]
rmse_elasticnet <- sqrt(mean((y - predictions_elasticnet)^2))
r2_elasticnet <- 1 - (sum((y - predictions_elasticnet)^2) / sum((y - mean(y))^2))

cat("Elastic Net RMSE: $", round(rmse_elasticnet, 2), "\n", sep = "")
cat("Elastic Net R-squared:", round(r2_elasticnet, 4), "\n\n")
```

## Regularization Comparison

```{r}
regularization_results <- data.frame(
  Model = c("Ridge", "Lasso", "Elastic Net"),
  RMSE = c(rmse_ridge, rmse_lasso, rmse_elasticnet),
  Rsquared = c(r2_ridge, r2_lasso, r2_elasticnet),
  Lambda = c(ridge_cv$lambda.min, lasso_cv$lambda.min, elasticnet_cv$lambda.min)
)

cat("Regularization Methods Comparison:\n")
print(regularization_results)

# Plot CV curves
par(mfrow = c(1, 3))
plot(ridge_cv, main = "Ridge Regression CV")
plot(lasso_cv, main = "Lasso Regression CV")
plot(elasticnet_cv, main = "Elastic Net CV")
par(mfrow = c(1, 1))
```

# 5. Tree-Based Models

```{r}
cat("\n===== 5. TREE-BASED MODELS =====\n\n")

set.seed(123)

# Random Forest
cat("Training Random Forest (ntree=100)...\n")
model_rf <- randomForest(
  sale_price ~ .,
  data = model_df_prepared,
  ntree = 100,
  mtry = 5,
  importance = TRUE,
  verbose = FALSE
)

predictions_rf <- predict(model_rf, newdata = model_df_prepared)
rmse_rf <- sqrt(mean((model_df_prepared$sale_price - predictions_rf)^2))
r2_rf <- 1 - (sum((model_df_prepared$sale_price - predictions_rf)^2) /
              sum((model_df_prepared$sale_price - mean(model_df_prepared$sale_price))^2))

cat("Random Forest RMSE: $", round(rmse_rf, 2), "\n", sep = "")
cat("Random Forest R-squared:", round(r2_rf, 4), "\n\n")

# Gradient Boosting Machine
cat("Training Gradient Boosting Machine...\n")
model_gbm <- gbm(
  sale_price ~ .,
  data = model_df_prepared,
  distribution = "gaussian",
  n.trees = 100,
  interaction.depth = 5,
  shrinkage = 0.1,
  verbose = FALSE
)

predictions_gbm <- predict(model_gbm, newdata = model_df_prepared, n.trees = 100)
rmse_gbm <- sqrt(mean((model_df_prepared$sale_price - predictions_gbm)^2))
r2_gbm <- 1 - (sum((model_df_prepared$sale_price - predictions_gbm)^2) /
               sum((model_df_prepared$sale_price - mean(model_df_prepared$sale_price))^2))

cat("Gradient Boosting RMSE: $", round(rmse_gbm, 2), "\n", sep = "")
cat("Gradient Boosting R-squared:", round(r2_gbm, 4), "\n\n")

# Feature Importance - Random Forest
cat("Random Forest Feature Importance (Top 15):\n")
importance_df <- data.frame(
  Feature = rownames(model_rf$importance),
  Importance = model_rf$importance[, 1]
) %>%
  arrange(desc(Importance)) %>%
  head(15)

print(importance_df)

# Feature Importance - Gradient Boosting
cat("\n\nGradient Boosting Feature Importance (Top 15):\n")
gbm_importance <- summary(model_gbm, plotit = FALSE) %>%
  arrange(desc(rel.inf)) %>%
  head(15)

print(gbm_importance)
```

## Feature Importance Visualization

```{r}
p_rf_imp <- ggplot(head(importance_df, 10),
                   aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Random Forest: Top 10 Features",
       x = "Feature",
       y = "Importance")

p_gbm_imp <- ggplot(head(gbm_importance, 10),
                    aes(x = reorder(var, rel.inf), y = rel.inf)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(title = "Gradient Boosting: Top 10 Features",
       x = "Feature",
       y = "Relative Influence")

print(p_rf_imp / p_gbm_imp)
```

# Comprehensive Model Comparison

```{r}
cat("\n===== COMPREHENSIVE MODEL COMPARISON =====\n\n")

all_models_comparison <- data.frame(
  Model = c(
    "Baseline Linear Regression",
    "Log-Transformed Linear Regression",
    "Polynomial Linear Regression",
    "Ridge Regression",
    "Lasso Regression",
    "Elastic Net",
    "Random Forest",
    "Gradient Boosting"
  ),
  RMSE = c(
    187104.50,
    rmse_log_transformed,
    rmse_poly,
    rmse_ridge,
    rmse_lasso,
    rmse_elasticnet,
    rmse_rf,
    rmse_gbm
  ),
  Rsquared = c(
    0.6087,
    summary(model_log_lr)$r.squared,
    summary(model_poly)$r.squared,
    r2_ridge,
    r2_lasso,
    r2_elasticnet,
    r2_rf,
    r2_gbm
  ),
  Parameters = c(
    22,
    22,
    28,
    "Regularized",
    "Regularized",
    "Regularized",
    "100 trees",
    "100 trees"
  )
)

all_models_comparison <- all_models_comparison %>%
  mutate(
    RMSE_Improvement = 187104.50 - RMSE,
    Percent_Improvement = round((RMSE_Improvement / 187104.50) * 100, 2)
  ) %>%
  arrange(RMSE)

cat("All Models Performance Comparison (Ranked by RMSE):\n\n")
print(all_models_comparison)

# Visualization
p_rmse_comp <- ggplot(all_models_comparison,
                      aes(x = reorder(Model, -RMSE), y = RMSE, fill = Rsquared)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  geom_hline(yintercept = 187104.50, linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "Model Comparison: RMSE and R-squared",
    subtitle = "Red line = Baseline Linear Regression",
    x = "Model",
    y = "RMSE ($)",
    fill = "R-squared"
  )

print(p_rmse_comp)

# R-squared comparison
p_r2_comp <- ggplot(all_models_comparison,
                    aes(x = reorder(Model, Rsquared), y = Rsquared, fill = Rsquared)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightcoral", high = "darkgreen") +
  geom_hline(yintercept = 0.6087, linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "Model Comparison: R-squared Values",
    subtitle = "Red line = Baseline Linear Regression",
    x = "Model",
    y = "R-squared",
    fill = "R-squared"
  )

print(p_r2_comp)
```

# Summary and Recommendations

```{r}
cat("\n===== SUMMARY AND KEY FINDINGS =====\n\n")

best_model <- all_models_comparison %>% slice(1)
best_model_name <- best_model$Model

cat("BEST PERFORMING MODEL:", best_model_name, "\n")
cat("RMSE: $", round(best_model$RMSE, 2), "\n", sep = "")
cat("R-squared:", round(best_model$Rsquared, 4), "\n")
cat("Improvement over baseline:",
    round(best_model$RMSE_Improvement, 2),
    " (", best_model$Percent_Improvement, "%)\n\n", sep = "")

cat("KEY FINDINGS:\n\n")
cat("1. LOG-TRANSFORMATION:\n")
cat("   - Improves residual normality\n")
cat("   - More stable predictions for extreme values\n")
cat("   - RMSE improvement: ",
    round((187104.50 - rmse_log_transformed), 2), "\n\n", sep = "")

cat("2. POLYNOMIAL FEATURES:\n")
cat("   - Captures non-linear relationships\n")
cat("   - Square of year_built and quality_score are significant\n")
cat("   - RMSE improvement: ",
    round((187104.50 - rmse_poly), 2), "\n\n", sep = "")

cat("3. CROSS-VALIDATION:\n")
cat("   - Provides robust out-of-sample performance estimates\n")
cat("   - Baseline CV RMSE: $",
    round(model_baseline_cv$results$RMSE[1], 2), "\n\n", sep = "")

cat("4. REGULARIZATION:\n")
cat("   - Ridge/Lasso prevent overfitting\n")
cat("   - Best regularization RMSE: $",
    round(min(rmse_ridge, rmse_lasso, rmse_elasticnet), 2), "\n\n", sep = "")

cat("5. TREE-BASED MODELS:\n")
cat("   - Random Forest RMSE: $", round(rmse_rf, 2), "\n", sep = "")
cat("   - Gradient Boosting RMSE: $", round(rmse_gbm, 2), "\n\n", sep = "")

cat("RECOMMENDED NEXT STEPS:\n")
cat("- Deploy", best_model_name, "for production predictions\n")
cat("- Monitor performance on new data\n")
cat("- Consider ensemble methods combining best models\n")
cat("- Perform hyperparameter tuning on", best_model_name, "\n")
cat("- Validate predictions on hold-out test set\n")
}
```

# Export Best Model

```{r}
# Save all comparison results
results_export <- all_models_comparison %>%
  select(Model, RMSE, Rsquared, RMSE_Improvement, Percent_Improvement)

write_csv(results_export, "models/model_comparison_results.csv")

# Export best models for deployment
saveRDS(model_log_lr, "models/log_transformed_lr_model.rds")
saveRDS(model_poly, "models/polynomial_lr_model.rds")
saveRDS(model_rf, "models/random_forest_model.rds")
saveRDS(model_gbm, "models/gradient_boosting_model.rds")
saveRDS(ridge_cv, "models/ridge_regression_model.rds")
saveRDS(lasso_cv, "models/lasso_regression_model.rds")

cat("\nAll models and results exported successfully!\n")
cat("Location: models/ directory\n")
cat("Summary: models/model_comparison_results.csv\n")
```
