---
title: "Regression Analysis with Transformations & Outlier Detection"
author: "Data Analysis Team"
date: "`r Sys.Date()`"
format: html
editor: visual
execute:
  warning: false
  message: false
---

# Advanced Regression Analysis: Transformations & Outlier Detection

This document implements focused regression analysis techniques including:

1. Multiple target variable transformations (Log, Square Root, Box-Cox, Reciprocal)
2. Outlier detection and analysis
3. Polynomial feature engineering for non-linear relationships
4. 10-fold cross-validation for robust evaluation
5. Regularization methods (Ridge/Lasso) to prevent overfitting

# Setup and Libraries

```{r}
load_packages <- function(packages) {
  lapply(packages, function(pkg) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, repos = "https://cloud.r-project.org")
      library(pkg, character.only = TRUE)
    }
  })
}

load_packages(c(
  "readr",
  "dplyr",
  "ggplot2",
  "caret",
  "glmnet",
  "tidyr",
  "broom",
  "scales",
  "patchwork",
  "gridExtra",
  "stringr",
  "MASS"
))
```

# Load Data

```{r}
app_model_df <- read_csv("data/house_price_model_data.csv")

cat("Dataset loaded successfully\n")
cat("Observations:", nrow(app_model_df), "\n")
cat("Features:", ncol(app_model_df), "\n")
```

# Prepare Data for Advanced Modeling

```{r}
# Prepare dataset by removing location_cluster
model_df_prepared <- app_model_df[, names(app_model_df) != "location_cluster"]

# Identify boolean features
boolean_features <- names(model_df_prepared)[sapply(model_df_prepared, is.logical)]

if (length(boolean_features) > 0) {
  for (col in boolean_features) {
    model_df_prepared[[col]] <- as.numeric(model_df_prepared[[col]]) - 1
  }
}

cat("Data prepared. Total features:", ncol(model_df_prepared) - 1, "\n")
```

# 1. Target Variable Transformations Analysis

```{r}
cat("===== 1. TARGET VARIABLE TRANSFORMATIONS =====\n\n")

# Helper function for skewness - MUST BE DEFINED FIRST
skewness <- function(x) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  n <- length(x)
  ((sum((x - m)^3, na.rm = TRUE) / n) / (s^3))
}

# Create transformation dataset using base R
model_df_transforms <- data.frame(model_df_prepared)
model_df_transforms$log_sale_price <- log(model_df_transforms$sale_price)
model_df_transforms$sqrt_sale_price <- sqrt(model_df_transforms$sale_price)
model_df_transforms$reciprocal_sale_price <- 1 / model_df_transforms$sale_price

# Calculate Box-Cox transformation lambda
boxcox_result <- boxcox(sale_price ~ 1, data = model_df_prepared, lambda = seq(-0.5, 1, 0.1))
lambda_optimal <- boxcox_result$x[which.max(boxcox_result$y)]

model_df_transforms$boxcox_sale_price <- if (abs(lambda_optimal) < 0.001) {
  log(model_df_prepared$sale_price)
} else {
  ((model_df_prepared$sale_price ^ lambda_optimal) - 1) / lambda_optimal
}

cat("Transformation Methods Applied:\n\n")

cat("1. ORIGINAL (No Transformation):\n")
cat("   Mean: $", round(mean(model_df_prepared$sale_price), 2), "\n", sep = "")
cat("   Median: $", round(median(model_df_prepared$sale_price), 2), "\n", sep = "")
cat("   Skewness:", round(skewness(model_df_prepared$sale_price), 4), "\n\n")

cat("2. LOG TRANSFORMATION: log(price)\n")
cat("   Mean:", round(mean(model_df_transforms$log_sale_price), 4), "\n")
cat("   Median:", round(median(model_df_transforms$log_sale_price), 4), "\n")
cat("   Skewness:", round(skewness(model_df_transforms$log_sale_price), 4), "\n\n")

cat("3. SQUARE ROOT TRANSFORMATION: sqrt(price)\n")
cat("   Mean:", round(mean(model_df_transforms$sqrt_sale_price), 2), "\n")
cat("   Median:", round(median(model_df_transforms$sqrt_sale_price), 2), "\n")
cat("   Skewness:", round(skewness(model_df_transforms$sqrt_sale_price), 4), "\n\n")

cat("4. RECIPROCAL TRANSFORMATION: 1/price\n")
cat("   Mean:", round(mean(model_df_transforms$reciprocal_sale_price), 6), "\n")
cat("   Median:", round(median(model_df_transforms$reciprocal_sale_price), 6), "\n")
cat("   Skewness:", round(skewness(model_df_transforms$reciprocal_sale_price), 4), "\n\n")

cat("5. BOX-COX TRANSFORMATION: lambda =", round(lambda_optimal, 4), "\n")
cat("   Mean:", round(mean(model_df_transforms$boxcox_sale_price), 2), "\n")
cat("   Median:", round(median(model_df_transforms$boxcox_sale_price), 2), "\n")
cat("   Skewness:", round(skewness(model_df_transforms$boxcox_sale_price), 4), "\n\n")

# Create visualization comparing distributions
p1 <- ggplot(model_df_transforms, aes(x = sale_price)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  labs(title = "Original Price Distribution",
       x = "Sale Price ($)",
       y = "Frequency") +
  scale_x_continuous(labels = scales::dollar)

p2 <- ggplot(model_df_transforms, aes(x = log_sale_price)) +
  geom_histogram(bins = 50, fill = "darkgreen", color = "black") +
  labs(title = "Log-Transformed Distribution",
       x = "Log(Sale Price)",
       y = "Frequency")

p3 <- ggplot(model_df_transforms, aes(x = sqrt_sale_price)) +
  geom_histogram(bins = 50, fill = "coral", color = "black") +
  labs(title = "Square Root Transformation",
       x = "sqrt(Sale Price)",
       y = "Frequency")

p4 <- ggplot(model_df_transforms, aes(x = boxcox_sale_price)) +
  geom_histogram(bins = 50, fill = "purple", color = "black") +
  labs(title = "Box-Cox Transformation",
       x = "Box-Cox(Sale Price)",
       y = "Frequency")

print((p1 | p2) / (p3 | p4))
```

## Build Linear Regression Models with Different Transformations

```{r}
cat("\nBuilding and comparing regression models with different transformations...\n\n")

# Model 1: Log-Transformed
model_log <- lm(log_sale_price ~ . -sale_price -sqrt_sale_price -reciprocal_sale_price -boxcox_sale_price,
                data = model_df_transforms)
pred_log <- exp(predict(model_log, newdata = model_df_transforms))
rmse_log <- sqrt(mean((model_df_prepared$sale_price - pred_log)^2))
r2_log <- summary(model_log)$r.squared

# Model 2: Square Root Transformed
model_sqrt <- lm(sqrt_sale_price ~ . -sale_price -log_sale_price -reciprocal_sale_price -boxcox_sale_price,
                 data = model_df_transforms)
pred_sqrt <- predict(model_sqrt, newdata = model_df_transforms) ^ 2
rmse_sqrt <- sqrt(mean((model_df_prepared$sale_price - pred_sqrt)^2))
r2_sqrt <- summary(model_sqrt)$r.squared

# Model 3: Box-Cox Transformed
model_boxcox <- lm(boxcox_sale_price ~ . -sale_price -log_sale_price -sqrt_sale_price -reciprocal_sale_price,
                   data = model_df_transforms)
# Inverse Box-Cox transformation
if (abs(lambda_optimal) < 0.001) {
  pred_boxcox <- exp(predict(model_boxcox, newdata = model_df_transforms))
} else {
  pred_boxcox <- (predict(model_boxcox, newdata = model_df_transforms) * lambda_optimal + 1) ^ (1 / lambda_optimal)
}
rmse_boxcox <- sqrt(mean((model_df_prepared$sale_price - pred_boxcox)^2))
r2_boxcox <- summary(model_boxcox)$r.squared

cat("Log-Transformed Model:\n")
cat("  R-squared:", round(r2_log, 4), "\n")
cat("  RMSE: $", round(rmse_log, 2), "\n\n", sep = "")

cat("Square Root Transformed Model:\n")
cat("  R-squared:", round(r2_sqrt, 4), "\n")
cat("  RMSE: $", round(rmse_sqrt, 2), "\n\n", sep = "")

cat("Box-Cox Transformed Model (lambda =", round(lambda_optimal, 4), "):\n")
cat("  R-squared:", round(r2_boxcox, 4), "\n")
cat("  RMSE: $", round(rmse_boxcox, 2), "\n\n", sep = "")

cat("Baseline (Original Scale): RMSE = $187,104.50, R-squared = 0.6087\n")
```

## Residual Normality Comparison

```{r}
residuals_comparison <- data.frame(
  Transformation = rep(c("Log", "Square Root", "Box-Cox"), each = nrow(model_df_transforms)),
  Residuals = c(
    residuals(model_log),
    residuals(model_sqrt),
    residuals(model_boxcox)
  )
)

p_qq_comp <- ggplot(residuals_comparison, aes(sample = Residuals)) +
  stat_qq(alpha = 0.3) +
  stat_qq_line(color = "red") +
  facet_wrap(~Transformation) +
  labs(title = "Q-Q Plots: Residual Normality by Transformation Method",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles")

print(p_qq_comp)
```

# 2. Polynomial Feature Engineering

```{r}
cat("\n===== 2. POLYNOMIAL FEATURE ENGINEERING =====\n\n")

model_df_poly <- model_df_prepared %>%
  mutate(
    square_feet_sq = square_feet^2,
    land_net_square_feet_sq = land_net_square_feet^2,
    bathrooms_sq = bathrooms^2,
    bedrooms_sq = bedrooms^2,
    year_built_sq = year_built^2,
    quality_score_sq = quality_score^2
  )

cat("Polynomial features created:\n")
cat("- square_feet_sq\n")
cat("- land_net_square_feet_sq\n")
cat("- bathrooms_sq\n")
cat("- bedrooms_sq\n")
cat("- year_built_sq\n")
cat("- quality_score_sq\n\n")

cat("New feature count:", ncol(model_df_poly) - 1, "\n")

# Build polynomial model
model_poly <- lm(sale_price ~ ., data = model_df_poly)

cat("\nPolynomial Model Summary:\n")
print(summary(model_poly))

cat("\n\nPolynomial Model Performance:\n")
cat("R-squared:", round(summary(model_poly)$r.squared, 4), "\n")
cat("Adjusted R-squared:", round(summary(model_poly)$adj.r.squared, 4), "\n")

predictions_poly <- predict(model_poly, newdata = model_df_poly)
residuals_poly <- model_df_poly$sale_price - predictions_poly
rmse_poly <- sqrt(mean(residuals_poly^2))
cat("RMSE: $", round(rmse_poly, 2), "\n", sep = "")
cat("Improvement vs baseline:", round(187104.50 - rmse_poly, 2), "\n\n")
```

## Comparison of Polynomial Terms Significance

```{r}
poly_coefs <- tidy(model_poly) %>%
  filter(str_detect(term, "_sq$")) %>%
  arrange(p.value)

cat("Polynomial Terms Significance (ordered by p-value):\n")
print(poly_coefs %>% select(term, estimate, p.value, statistic))
```

# 3. 10-Fold Cross-Validation

```{r}
cat("\n===== 3. 10-FOLD CROSS-VALIDATION =====\n\n")

set.seed(123)

# Define cross-validation setup
ctrl <- trainControl(
  method = "cv",
  number = 10,
  verboseIter = FALSE
)

# Model 1: Baseline Linear Regression
cat("Training Baseline Linear Regression with 10-fold CV...\n")
model_baseline_cv <- train(
  sale_price ~ .,
  data = model_df_prepared,
  method = "lm",
  trControl = ctrl
)

cat("\nBaseline Linear Regression - 10-Fold CV Results:\n")
print(model_baseline_cv)

# Model 2: Polynomial Linear Regression
cat("\n\nTraining Polynomial Linear Regression with 10-fold CV...\n")
model_poly_cv <- train(
  sale_price ~ .,
  data = model_df_poly,
  method = "lm",
  trControl = ctrl,
  preProcess = c("center", "scale")
)

cat("\nPolynomial Linear Regression - 10-Fold CV Results:\n")
print(model_poly_cv)
```

## Cross-Validation Results Comparison

```{r}
cv_results <- data.frame(
  Model = c("Baseline LR", "Polynomial LR"),
  RMSE_Mean = c(
    model_baseline_cv$results$RMSE[1],
    model_poly_cv$results$RMSE[1]
  ),
  RMSE_SD = c(
    model_baseline_cv$results$RMSESD[1],
    model_poly_cv$results$RMSESD[1]
  ),
  Rsquared_Mean = c(
    model_baseline_cv$results$Rsquared[1],
    model_poly_cv$results$Rsquared[1]
  )
)

cat("\n10-Fold Cross-Validation Summary:\n")
print(cv_results)

# Visualize CV results
cv_comparison <- data.frame(
  Model = rep(c("Baseline", "Polynomial"), each = 10),
  RMSE = c(
    model_baseline_cv$resample$RMSE,
    model_poly_cv$resample$RMSE
  )
)

p_cv <- ggplot(cv_comparison, aes(x = Model, y = RMSE, fill = Model)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "10-Fold Cross-Validation RMSE Comparison",
       y = "RMSE ($)") +
  theme(legend.position = "bottom")

print(p_cv)
```

# 4. Regularization Methods (Ridge and Lasso)

```{r}
cat("\n===== 4. REGULARIZATION METHODS: RIDGE & LASSO =====\n\n")

# Prepare data for glmnet (matrix format)
X <- model_df_prepared %>%
  select(-sale_price) %>%
  as.matrix()

y <- model_df_prepared$sale_price

# Ridge Regression
cat("Training Ridge Regression with CV...\n")
ridge_cv <- cv.glmnet(
  X, y,
  alpha = 0,
  nfolds = 10,
  type.measure = "mse"
)

cat("Ridge Regression - Best Lambda:", ridge_cv$lambda.min, "\n")

predictions_ridge <- predict(ridge_cv, newx = X, s = "lambda.min")[, 1]
rmse_ridge <- sqrt(mean((y - predictions_ridge)^2))
r2_ridge <- 1 - (sum((y - predictions_ridge)^2) / sum((y - mean(y))^2))

cat("Ridge RMSE: $", round(rmse_ridge, 2), "\n", sep = "")
cat("Ridge R-squared:", round(r2_ridge, 4), "\n\n")

# Lasso Regression
cat("Training Lasso Regression with CV...\n")
lasso_cv <- cv.glmnet(
  X, y,
  alpha = 1,
  nfolds = 10,
  type.measure = "mse"
)

cat("Lasso Regression - Best Lambda:", lasso_cv$lambda.min, "\n")

predictions_lasso <- predict(lasso_cv, newx = X, s = "lambda.min")[, 1]
rmse_lasso <- sqrt(mean((y - predictions_lasso)^2))
r2_lasso <- 1 - (sum((y - predictions_lasso)^2) / sum((y - mean(y))^2))

cat("Lasso RMSE: $", round(rmse_lasso, 2), "\n", sep = "")
cat("Lasso R-squared:", round(r2_lasso, 4), "\n\n")

# Elastic Net (combination)
cat("Training Elastic Net (alpha=0.5) with CV...\n")
elasticnet_cv <- cv.glmnet(
  X, y,
  alpha = 0.5,
  nfolds = 10,
  type.measure = "mse"
)

cat("Elastic Net - Best Lambda:", elasticnet_cv$lambda.min, "\n")

predictions_elasticnet <- predict(elasticnet_cv, newx = X, s = "lambda.min")[, 1]
rmse_elasticnet <- sqrt(mean((y - predictions_elasticnet)^2))
r2_elasticnet <- 1 - (sum((y - predictions_elasticnet)^2) / sum((y - mean(y))^2))

cat("Elastic Net RMSE: $", round(rmse_elasticnet, 2), "\n", sep = "")
cat("Elastic Net R-squared:", round(r2_elasticnet, 4), "\n\n")
```

## Regularization Comparison

```{r}
regularization_results <- data.frame(
  Model = c("Ridge", "Lasso", "Elastic Net"),
  RMSE = c(rmse_ridge, rmse_lasso, rmse_elasticnet),
  Rsquared = c(r2_ridge, r2_lasso, r2_elasticnet),
  Lambda = c(ridge_cv$lambda.min, lasso_cv$lambda.min, elasticnet_cv$lambda.min)
)

cat("Regularization Methods Comparison:\n")
print(regularization_results)

# Plot CV curves
par(mfrow = c(1, 3))
plot(ridge_cv, main = "Ridge Regression CV")
plot(lasso_cv, main = "Lasso Regression CV")
plot(elasticnet_cv, main = "Elastic Net CV")
par(mfrow = c(1, 1))
```

# 5. Outlier Detection and Analysis

```{r}
cat("\n===== 5. OUTLIER DETECTION =====\n\n")

# Method 1: IQR Method
cat("Outlier Detection - Method 1: Interquartile Range (IQR)\n\n")

Q1 <- quantile(model_df_prepared$sale_price, 0.25)
Q3 <- quantile(model_df_prepared$sale_price, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

outliers_iqr <- model_df_prepared %>%
  mutate(is_outlier_iqr = sale_price < lower_bound | sale_price > upper_bound)

n_outliers_iqr <- sum(outliers_iqr$is_outlier_iqr)
pct_outliers_iqr <- (n_outliers_iqr / nrow(model_df_prepared)) * 100

cat("IQR Method Results:\n")
cat("Lower Bound: $", round(lower_bound, 2), "\n", sep = "")
cat("Upper Bound: $", round(upper_bound, 2), "\n", sep = "")
cat("Number of outliers:", n_outliers_iqr, "\n")
cat("Percentage of outliers:", round(pct_outliers_iqr, 2), "%\n\n")

# Method 2: Z-Score Method
cat("Outlier Detection - Method 2: Z-Score (|Z| > 3)\n\n")

z_scores <- scale(model_df_prepared$sale_price)[, 1]
outliers_zscore <- abs(z_scores) > 3

n_outliers_zscore <- sum(outliers_zscore)
pct_outliers_zscore <- (n_outliers_zscore / nrow(model_df_prepared)) * 100

cat("Z-Score Method Results:\n")
cat("Number of outliers (|Z| > 3):", n_outliers_zscore, "\n")
cat("Percentage of outliers:", round(pct_outliers_zscore, 2), "%\n\n")

# Method 3: Modified Z-Score (using MAD - Median Absolute Deviation)
cat("Outlier Detection - Method 3: Modified Z-Score (MAD)\n\n")

median_price <- median(model_df_prepared$sale_price)
mad <- median(abs(model_df_prepared$sale_price - median_price))
modified_z <- 0.6745 * (model_df_prepared$sale_price - median_price) / mad
outliers_mad <- abs(modified_z) > 3.5

n_outliers_mad <- sum(outliers_mad)
pct_outliers_mad <- (n_outliers_mad / nrow(model_df_prepared)) * 100

cat("Modified Z-Score (MAD) Method Results:\n")
cat("Number of outliers:", n_outliers_mad, "\n")
cat("Percentage of outliers:", round(pct_outliers_mad, 2), "%\n\n")

# Visualization of outliers
outlier_data <- model_df_prepared %>%
  mutate(
    outlier_iqr = if_else(is_outlier_iqr, "Outlier", "Normal"),
    outlier_mad = if_else(outliers_mad, "Outlier", "Normal")
  )

p_outliers_iqr <- ggplot(outlier_data, aes(x = "", y = sale_price, fill = outlier_iqr)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.1, size = 1) +
  scale_y_continuous(labels = scales::dollar) +
  labs(
    title = "Outlier Detection: IQR Method",
    y = "Sale Price ($)",
    fill = "Status"
  ) +
  theme(axis.title.x = element_blank(), axis.text.x = element_blank())

p_outliers_dist <- ggplot(outlier_data, aes(x = sale_price, fill = outlier_mad)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title = "Price Distribution with MAD Outliers",
    x = "Sale Price ($)",
    y = "Frequency",
    fill = "Status"
  )

print(p_outliers_iqr / p_outliers_dist)
```

## Summary of Outlier Methods

```{r}
outlier_summary <- data.frame(
  Method = c("IQR Method", "Z-Score (|Z| > 3)", "Modified Z-Score (MAD)"),
  Number_Outliers = c(n_outliers_iqr, n_outliers_zscore, n_outliers_mad),
  Percentage = c(pct_outliers_iqr, pct_outliers_zscore, pct_outliers_mad)
)

cat("\nOutlier Detection Methods Comparison:\n")
print(outlier_summary)

cat("\nInterpretation:\n")
cat("- IQR Method: Conservative, detects extreme values beyond 1.5*IQR\n")
cat("- Z-Score: Less robust to skewness, sensitive to distribution shape\n")
cat("- Modified Z-Score (MAD): More robust, less affected by outliers in calculation\n\n")

# Compare outlier sets
cat("Outlier Agreement Analysis:\n")
combined_outliers <- data.frame(
  IQR = outliers_iqr$is_outlier_iqr,
  ZScore = outliers_zscore,
  MAD = outliers_mad
) %>%
  mutate(Agreement = rowSums(across(everything())) >= 2)

n_all_three <- sum(combined_outliers$IQR & combined_outliers$ZScore & combined_outliers$MAD)
n_at_least_two <- sum(combined_outliers$Agreement)

cat("Outliers detected by all 3 methods:", n_all_three, "\n")
cat("Outliers detected by at least 2 methods:", n_at_least_two, "\n")
```

# Comprehensive Model Comparison

```{r}
cat("\n===== COMPREHENSIVE REGRESSION MODEL COMPARISON =====\n\n")

all_models_comparison <- data.frame(
  Model = c(
    "Baseline Linear Regression",
    "Log-Transformed Linear Regression",
    "Square Root Transformed",
    "Box-Cox Transformed",
    "Polynomial Linear Regression",
    "Ridge Regression",
    "Lasso Regression",
    "Elastic Net"
  ),
  RMSE = c(
    187104.50,
    rmse_log,
    rmse_sqrt,
    rmse_boxcox,
    rmse_poly,
    rmse_ridge,
    rmse_lasso,
    rmse_elasticnet
  ),
  Rsquared = c(
    0.6087,
    r2_log,
    r2_sqrt,
    r2_boxcox,
    summary(model_poly)$r.squared,
    r2_ridge,
    r2_lasso,
    r2_elasticnet
  ),
  Model_Type = c(
    "Original Scale",
    "Log Transform",
    "Square Root Transform",
    "Box-Cox Transform",
    "Polynomial Features",
    "Regularization",
    "Regularization",
    "Regularization"
  )
)

all_models_comparison <- all_models_comparison %>%
  mutate(
    RMSE_Improvement = 187104.50 - RMSE,
    Percent_Improvement = round((RMSE_Improvement / 187104.50) * 100, 2)
  ) %>%
  arrange(RMSE)

cat("All Models Performance Comparison (Ranked by RMSE):\n\n")
print(all_models_comparison)

# Visualization
p_rmse_comp <- ggplot(all_models_comparison,
                      aes(x = reorder(Model, -RMSE), y = RMSE, fill = Rsquared)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  geom_hline(yintercept = 187104.50, linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "Model Comparison: RMSE and R-squared",
    subtitle = "Red line = Baseline Linear Regression",
    x = "Model",
    y = "RMSE ($)",
    fill = "R-squared"
  )

print(p_rmse_comp)

# R-squared comparison
p_r2_comp <- ggplot(all_models_comparison,
                    aes(x = reorder(Model, Rsquared), y = Rsquared, fill = Rsquared)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightcoral", high = "darkgreen") +
  geom_hline(yintercept = 0.6087, linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "Model Comparison: R-squared Values",
    subtitle = "Red line = Baseline Linear Regression",
    x = "Model",
    y = "R-squared",
    fill = "R-squared"
  )

print(p_r2_comp)
```

# Summary and Recommendations

```{r}
cat("\n===== SUMMARY AND KEY FINDINGS =====\n\n")

best_model <- all_models_comparison %>% slice(1)
best_model_name <- best_model$Model

cat("BEST PERFORMING MODEL:", best_model_name, "\n")
cat("RMSE: $", round(best_model$RMSE, 2), "\n", sep = "")
cat("R-squared:", round(best_model$Rsquared, 4), "\n")
cat("Improvement over baseline:",
    round(best_model$RMSE_Improvement, 2),
    " (", best_model$Percent_Improvement, "%)\n\n", sep = "")

cat("KEY FINDINGS:\n\n")
cat("1. TARGET VARIABLE TRANSFORMATIONS:\n")
cat("   - Log-Transformation RMSE: $", round(rmse_log, 2), "\n", sep = "")
cat("     • Improves residual normality\n")
cat("     • Helps with heteroscedasticity\n")
cat("     • Better for skewed price distributions\n\n")

cat("   - Square Root Transformation RMSE: $", round(rmse_sqrt, 2), "\n", sep = "")
cat("     • Less aggressive than log\n")
cat("     • Maintains interpretability\n\n")

cat("   - Box-Cox Transformation RMSE: $", round(rmse_boxcox, 2), "\n", sep = "")
cat("     • Optimal lambda:", round(lambda_optimal, 4), "\n")
cat("     • Data-driven transformation\n")
cat("     • Theoretically optimal for normality\n\n")

cat("2. POLYNOMIAL FEATURES:\n")
cat("   - Captures non-linear relationships\n")
cat("   - RMSE: $", round(rmse_poly, 2), "\n", sep = "")
cat("   - Improvement: ", round(187104.50 - rmse_poly, 2), " dollars\n\n", sep = "")

cat("3. CROSS-VALIDATION RESULTS:\n")
cat("   - Baseline CV RMSE: $", round(model_baseline_cv$results$RMSE[1], 2), "\n", sep = "")
cat("   - Polynomial CV RMSE: $", round(model_poly_cv$results$RMSE[1], 2), "\n")
cat("   - Provides robust out-of-sample estimates\n\n")

cat("4. REGULARIZATION TECHNIQUES:\n")
cat("   - Ridge Regression RMSE: $", round(rmse_ridge, 2), "\n", sep = "")
cat("   - Lasso Regression RMSE: $", round(rmse_lasso, 2), "\n", sep = "")
cat("   - Elastic Net RMSE: $", round(rmse_elasticnet, 2), "\n")
cat("   - Prevent overfitting through penalty terms\n\n")

cat("5. OUTLIER DETECTION FINDINGS:\n")
cat("   - IQR Method identified: ", n_outliers_iqr, " outliers (",
    round(pct_outliers_iqr, 2), "%)\n", sep = "")
cat("   - Modified Z-Score identified: ", n_outliers_mad, " outliers (",
    round(pct_outliers_mad, 2), "%)\n", sep = "")
cat("   - Recommend reviewing extreme values\n\n")

cat("RECOMMENDED APPROACH:\n")
cat("- Log-transformation is recommended for this dataset\n")
cat("- Use 10-fold cross-validation for robust validation\n")
cat("- Ridge/Lasso regularization to prevent overfitting\n")
cat("- Monitor and potentially remove high-leverage outliers\n")
cat("- Combine transformation with polynomial features for improved fit\n")
}
```

# Export Models and Results

```{r}
# Create and save comparison results
results_export <- all_models_comparison %>%
  arrange(RMSE) %>%
  mutate(
    RMSE_Improvement = 187104.50 - RMSE,
    Percent_Improvement = round((RMSE_Improvement / 187104.50) * 100, 2)
  )

write_csv(results_export, "models/transformation_model_comparison.csv")

# Export transformation-based models
saveRDS(model_log, "models/log_transformed_regression.rds")
saveRDS(model_sqrt, "models/sqrt_transformed_regression.rds")
saveRDS(model_boxcox, "models/boxcox_transformed_regression.rds")
saveRDS(model_poly, "models/polynomial_regression.rds")

# Export regularization models
saveRDS(ridge_cv, "models/ridge_regression.rds")
saveRDS(lasso_cv, "models/lasso_regression.rds")
saveRDS(elasticnet_cv, "models/elasticnet_regression.rds")

# Save transformation metadata
transformation_metadata <- list(
  lambda_boxcox = lambda_optimal,
  outlier_methods = list(
    iqr_count = n_outliers_iqr,
    iqr_percentage = pct_outliers_iqr,
    mad_count = n_outliers_mad,
    mad_percentage = pct_outliers_mad
  ),
  baseline_rmse = 187104.50,
  baseline_rsquared = 0.6087
)

saveRDS(transformation_metadata, "models/transformation_metadata.rds")

cat("\n===== MODEL EXPORT SUMMARY =====\n\n")
cat("Transformation-Based Models Saved:\n")
cat("  1. log_transformed_regression.rds\n")
cat("  2. sqrt_transformed_regression.rds\n")
cat("  3. boxcox_transformed_regression.rds\n")
cat("  4. polynomial_regression.rds\n\n")

cat("Regularization Models Saved:\n")
cat("  5. ridge_regression.rds\n")
cat("  6. lasso_regression.rds\n")
cat("  7. elasticnet_regression.rds\n\n")

cat("Results and Metadata Saved:\n")
cat("  - transformation_model_comparison.csv\n")
cat("  - transformation_metadata.rds\n\n")

print(results_export %>% select(Model, RMSE, Rsquared, Percent_Improvement))
```
