---
title: "Transformations & Outlier Detection - House Price Prediction"
author: "Data Analysis Team"
date: "`r Sys.Date()`"
format: html
execute:
  warning: false
  message: false
---

# Regression Analysis: Target Variable Transformations & Outlier Detection

This analysis focuses on improving linear regression through transformations and outlier detection methods.

# Setup

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(MASS)
library(patchwork)
library(scales)

app_model_df <- read_csv("data/house_price_model_data.csv")
model_df <- app_model_df[, names(app_model_df) != "location_cluster"]

cat("Dataset loaded:", nrow(model_df), "observations,", ncol(model_df), "features\n")
```

# 1. Analyze Original Price Distribution

```{r}
cat("\n===== ORIGINAL PRICE DISTRIBUTION =====\n\n")

cat("Summary Statistics:\n")
cat("Mean: $", round(mean(model_df$sale_price), 2), "\n", sep = "")
cat("Median: $", round(median(model_df$sale_price), 2), "\n", sep = "")
cat("Std Dev: $", round(sd(model_df$sale_price), 2), "\n", sep = "")
cat("Min: $", round(min(model_df$sale_price), 2), "\n", sep = "")
cat("Max: $", round(max(model_df$sale_price), 2), "\n\n", sep = "")

# Calculate skewness
mean_price <- mean(model_df$sale_price)
sd_price <- sd(model_df$sale_price)
skew_original <- mean((model_df$sale_price - mean_price)^3) / (sd_price^3)

cat("Skewness:", round(skew_original, 4), "\n")
cat("(Positive skewness = right-skewed, has long tail to right)\n\n")

p_original <- ggplot(model_df, aes(x = sale_price)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Original Sale Price Distribution",
       x = "Sale Price ($)",
       y = "Frequency") +
  scale_x_continuous(labels = dollar)

print(p_original)
```

# 2. Target Variable Transformations

```{r}
cat("\n===== TARGET VARIABLE TRANSFORMATIONS =====\n\n")

# Create transformed versions
trans_data <- data.frame(
  sale_price = model_df$sale_price,
  log_price = log(model_df$sale_price),
  sqrt_price = sqrt(model_df$sale_price)
)

# Box-Cox transformation
cat("Computing Box-Cox transformation...\n")
bc_fit <- boxcox(sale_price ~ 1, data = model_df, lambda = seq(-0.5, 1, 0.1), plotit = FALSE)
lambda_opt <- bc_fit$x[which.max(bc_fit$y)]

if (abs(lambda_opt) < 0.001) {
  trans_data$boxcox_price <- log(model_df$sale_price)
} else {
  trans_data$boxcox_price <- ((model_df$sale_price ^ lambda_opt) - 1) / lambda_opt
}

cat("Optimal Box-Cox lambda:", round(lambda_opt, 4), "\n\n")

# Calculate statistics for each transformation
transformations <- list(
  list(name = "Original", data = trans_data$sale_price),
  list(name = "Log", data = trans_data$log_price),
  list(name = "Square Root", data = trans_data$sqrt_price),
  list(name = "Box-Cox", data = trans_data$boxcox_price)
)

results <- data.frame()

for (trans in transformations) {
  mean_val <- mean(trans$data, na.rm = TRUE)
  sd_val <- sd(trans$data, na.rm = TRUE)
  skew_val <- mean((trans$data - mean_val)^3, na.rm = TRUE) / (sd_val^3)

  results <- rbind(results, data.frame(
    Transformation = trans$name,
    Mean = round(mean_val, 2),
    Std_Dev = round(sd_val, 2),
    Skewness = round(skew_val, 4)
  ))
}

cat("Transformation Comparison:\n")
print(results)

cat("\nInterpretation:\n")
cat("- Lower absolute skewness = better normality\n")
cat("- Log transformation typically works best for right-skewed price data\n")
cat("- Box-Cox gives optimal data-driven transformation\n\n")
```

## Visualization of Transformations

```{r}
p1 <- ggplot(trans_data, aes(x = sale_price)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Original", x = "Price ($)", y = "Frequency") +
  scale_x_continuous(labels = dollar)

p2 <- ggplot(trans_data, aes(x = log_price)) +
  geom_histogram(bins = 60, fill = "darkgreen", color = "black", alpha = 0.7) +
  labs(title = "Log(Price)", x = "Log(Price)", y = "Frequency")

p3 <- ggplot(trans_data, aes(x = sqrt_price)) +
  geom_histogram(bins = 60, fill = "coral", color = "black", alpha = 0.7) +
  labs(title = "Sqrt(Price)", x = "Sqrt(Price)", y = "Frequency")

p4 <- ggplot(trans_data, aes(x = boxcox_price)) +
  geom_histogram(bins = 60, fill = "purple", color = "black", alpha = 0.7) +
  labs(title = paste("Box-Cox (λ =", round(lambda_opt, 3), ")"),
       x = "Transformed Price", y = "Frequency")

print((p1 | p2) / (p3 | p4))
```

# 3. Outlier Detection

```{r}
cat("\n===== OUTLIER DETECTION =====\n\n")

price <- model_df$sale_price

# Method 1: IQR Method
cat("METHOD 1: Interquartile Range (IQR)\n")
cat("Rule: Outliers are values outside [Q1 - 1.5*IQR, Q3 + 1.5*IQR]\n\n")

Q1 <- quantile(price, 0.25)
Q3 <- quantile(price, 0.75)
IQR <- Q3 - Q1
lower_iqr <- Q1 - 1.5 * IQR
upper_iqr <- Q3 + 1.5 * IQR

outliers_iqr <- price < lower_iqr | price > upper_iqr
n_outliers_iqr <- sum(outliers_iqr)
pct_iqr <- (n_outliers_iqr / length(price)) * 100

cat("Lower Bound: $", round(lower_iqr, 0), "\n", sep = "")
cat("Upper Bound: $", round(upper_iqr, 0), "\n", sep = "")
cat("Outliers Detected:", n_outliers_iqr, "(", round(pct_iqr, 2), "%)\n\n")

# Method 2: Z-Score Method
cat("METHOD 2: Z-Score (|Z| > 3)\n")
cat("Rule: Values more than 3 standard deviations from mean\n\n")

z_scores <- (price - mean(price)) / sd(price)
outliers_z <- abs(z_scores) > 3
n_outliers_z <- sum(outliers_z)
pct_z <- (n_outliers_z / length(price)) * 100

cat("Outliers Detected:", n_outliers_z, "(", round(pct_z, 2), "%)\n\n")

# Method 3: Modified Z-Score (MAD - Median Absolute Deviation)
cat("METHOD 3: Modified Z-Score using MAD\n")
cat("Rule: |Modified Z| > 3.5, more robust to extreme values\n\n")

median_price <- median(price)
mad <- median(abs(price - median_price))
modified_z <- 0.6745 * (price - median_price) / mad
outliers_mad <- abs(modified_z) > 3.5
n_outliers_mad <- sum(outliers_mad)
pct_mad <- (n_outliers_mad / length(price)) * 100

cat("Outliers Detected:", n_outliers_mad, "(", round(pct_mad, 2), "%)\n\n")

# Summary comparison
outlier_summary <- data.frame(
  Method = c("IQR", "Z-Score", "Modified Z-Score"),
  Count = c(n_outliers_iqr, n_outliers_z, n_outliers_mad),
  Percentage = c(round(pct_iqr, 2), round(pct_z, 2), round(pct_mad, 2))
)

cat("Summary Comparison:\n")
print(outlier_summary)
```

## Outlier Visualization

```{r}
outlier_df <- data.frame(
  price = price,
  outlier_iqr = outliers_iqr,
  outlier_mad = outliers_mad
)

p_iqr <- ggplot(outlier_df, aes(x = "", y = price, fill = outlier_iqr)) +
  geom_boxplot(width = 0.3) +
  scale_y_continuous(labels = dollar) +
  scale_fill_manual(values = c("FALSE" = "lightblue", "TRUE" = "red")) +
  labs(title = "IQR Method", y = "Price ($)", fill = "Outlier") +
  theme(axis.title.x = element_blank(), axis.text.x = element_blank())

p_dist <- ggplot(outlier_df, aes(x = price, fill = outlier_mad)) +
  geom_histogram(bins = 60, alpha = 0.7, color = "black") +
  scale_x_continuous(labels = dollar) +
  scale_fill_manual(values = c("FALSE" = "lightblue", "TRUE" = "red")) +
  labs(title = "Modified Z-Score Distribution", x = "Price ($)", y = "Frequency", fill = "Outlier")

print(p_iqr / p_dist)
```

# 4. Impact of Removing Outliers

```{r}
cat("\n===== IMPACT OF OUTLIER REMOVAL =====\n\n")

# Create dataset without outliers
model_df_no_outliers <- model_df[!outliers_iqr, ]

cat("Original Dataset:\n")
cat("  Observations:", nrow(model_df), "\n")
cat("  Mean Price: $", round(mean(model_df$sale_price), 0), "\n", sep = "")
cat("  Std Dev: $", round(sd(model_df$sale_price), 0), "\n\n", sep = "")

cat("After Removing IQR Outliers:\n")
cat("  Observations:", nrow(model_df_no_outliers), "\n")
cat("  Removed:", nrow(model_df) - nrow(model_df_no_outliers), "rows\n")
cat("  Mean Price: $", round(mean(model_df_no_outliers$sale_price), 0), "\n", sep = "")
cat("  Std Dev: $", round(sd(model_df_no_outliers$sale_price), 0), "\n\n", sep = "")

cat("Change in Standard Deviation:",
    round((sd(model_df$sale_price) - sd(model_df_no_outliers$sale_price)) / sd(model_df$sale_price) * 100, 2),
    "%\n")
```

# 5. Build Linear Regression Models

```{r}
cat("\n===== LINEAR REGRESSION MODELS =====\n\n")

# Model 1: Baseline (Original Scale)
cat("Model 1: Baseline Linear Regression (Original Scale)\n")
model_baseline <- lm(sale_price ~ ., data = model_df[, 1:15])
pred_baseline <- predict(model_baseline, model_df[, 1:15])
rmse_baseline <- sqrt(mean((model_df$sale_price - pred_baseline)^2))
r2_baseline <- summary(model_baseline)$r.squared

cat("  R-squared:", round(r2_baseline, 4), "\n")
cat("  RMSE: $", round(rmse_baseline, 0), "\n\n", sep = "")

# Model 2: Log-Transformed
cat("Model 2: Log-Transformed Linear Regression\n")
model_log_df <- data.frame(log_price = log(model_df$sale_price), model_df[, 2:15])
model_log <- lm(log_price ~ ., data = model_log_df)
pred_log <- exp(predict(model_log, model_log_df))
rmse_log <- sqrt(mean((model_df$sale_price - pred_log)^2))
r2_log <- summary(model_log)$r.squared

cat("  R-squared:", round(r2_log, 4), "\n")
cat("  RMSE: $", round(rmse_log, 0), "\n")
cat("  Improvement: $", round(rmse_baseline - rmse_log, 0), "\n\n", sep = "")

# Model 3: Without Outliers
cat("Model 3: Linear Regression (Outliers Removed)\n")
model_no_out <- lm(sale_price ~ ., data = model_df_no_outliers[, 1:15])
pred_no_out <- predict(model_no_out, model_df_no_outliers[, 1:15])
rmse_no_out <- sqrt(mean((model_df_no_outliers$sale_price - pred_no_out)^2))
r2_no_out <- summary(model_no_out)$r.squared

cat("  R-squared:", round(r2_no_out, 4), "\n")
cat("  RMSE: $", round(rmse_no_out, 0), "\n")
cat("  Improvement: $", round(rmse_baseline - rmse_no_out, 0), "\n\n", sep = "")
```

# Summary & Recommendations

```{r}
cat("\n===== SUMMARY & RECOMMENDATIONS =====\n\n")

cat("KEY FINDINGS:\n\n")

cat("1. TRANSFORMATIONS:\n")
cat("   - Original data shows right-skewed distribution\n")
cat("   - Log transformation significantly improves normality\n")
cat("   - Box-Cox optimal lambda:", round(lambda_opt, 4), "\n")
cat("   - Log transformation RMSE improvement: $", round(rmse_baseline - rmse_log, 0), "\n\n", sep = "")

cat("2. OUTLIERS:\n")
cat("   - IQR method identifies", n_outliers_iqr, "outliers (", round(pct_iqr, 2), "%)\n")
cat("   - High-value properties detected as outliers\n")
cat("   - Removing outliers improves std dev but changes model fit\n\n")

cat("3. MODEL PERFORMANCE:\n")
cat("   - Baseline RMSE: $", round(rmse_baseline, 0), "\n", sep = "")
cat("   - Log-transformed: $", round(rmse_log, 0), "\n", sep = "")
cat("   - Without outliers: $", round(rmse_no_out, 0), "\n\n", sep = "")

cat("RECOMMENDATIONS:\n")
cat("- Use LOG-TRANSFORMATION for better model assumptions\n")
cat("- Keep outliers for now (represent legitimate high-value properties)\n")
cat("- Apply next: polynomial features, cross-validation, regularization\n")
cat("- Monitor outlier predictions separately if needed\n")
```
