---
title: "KNN model"
format:
  html:
    toc: true
    number-sections: true
---

# Introduction

This document provides a detailed interpretation of the KNN model applied to predict residential property prices in Pierce County. The analysis builds upon the data prepared during the exploratory phase and focuses on identifying the key drivers of house prices and assessing the predictive power of the resulting models.

# Step-by-Step Analysis

```{r}
#| label: setup
#| message: false
#| warning: false

library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(tidyr)
library(here)
library(dotenv)
library(caret)
library(tidyverse)
library(kknn)
options(scipen = 999)

# Load the prepared data
model_df <- read_csv(here::here("data", "house_price_model_data.csv"))
```


## Step 1: Data Preparation and Feature Selection

The analysis utilizes the `house_price_model_data.csv` dataset, which contains processed features including property characteristics and location data.

```{r}
#| label: data-summary
#| echo: false

cat("Total observations:", nrow(model_df), "\n")
cat("Total features available:", ncol(model_df) - 1, "\n")
```

The idea behind experimenting with KNN is to easily show the user why the given price
has been estimated through a clear reasoning (showing n most similar observations)
based on a very limited number of features (location and size, optionally condition).

Key features include:

- **Physical Characteristics**: `square_feet`, `bathrooms`, `bedrooms`
- **Quality Indicators**: `quality_score`, `has_pool_or_spa_detail`
- **Location**: `latitude`, `longitude`
- **Time**: `sale year`.

## Step 2: Model Development

### Minimal Model
We focus on the **Core Features Model** as it provides the most interpretable results
to the user with using only longitude, latitude and
the net square footage of properties.

```{r}
#| label: model-development

baseline_data <- model_df %>% 
  select(
    sale_price, 
    latitude, 
    longitude, 
    square_feet
  ) %>% 
  drop_na() # in case, but no NA

set.seed(222) 
train_index <- createDataPartition(
  baseline_data$sale_price, 
  p = 0.8, 
  list = FALSE
)

train_data_1 <- baseline_data[train_index, ]
test_data_1  <- baseline_data[-train_index, ]
```

We try for different values of k (2, 3, 4, 5). We decided these values as in a map
visualization (showing the closest observations to the user) more than 5 similar 
properties displayed would compromise the user experience.
```{r}

cv_control <- trainControl(
  method = "cv", 
  number = 3
)

k_grid <- data.frame(k = c(2, 3, 4, 5))

set.seed(222)
knn_baseline <- train(
  sale_price ~ latitude + longitude + square_feet,
  data = train_data_1,
  method = "knn",                  
  trControl = cv_control,
  preProcess = c("center", "scale"),
  tuneGrid = k_grid
)

print(knn_baseline)

```
```{r}
saveRDS(knn_baseline, here::here("models", "knn_baseline.rds"))
```

The optimal number of k is 5. This model has an RMSE of 247 381 dollars on validation
set. The R-squared is 0.3, which considering that we only used the property
coordinates and the net square footage, is not a bad result.
```{r}
# Predict on the test data
predictions_1 <- predict(knn_baseline, newdata = test_data_1)

# Calculate performance metrics (RMSE, R-squared, MAE)
evaluation_metrics_1 <- postResample(pred = predictions_1, obs = test_data_1$sale_price)
print(evaluation_metrics_1)

```
The RMSE is 234 085 dollars on the test set, consistent with the cross validation
so no over-fitting occurred.

```{r}
train_preds_1 <- predict(knn_baseline, newdata = train_data_1)

# Absolute Percentage Error (APE) for training data
train_ape_1 <- abs(train_preds_1 - train_data_1$sale_price) / train_data_1$sale_price

train_within_15_1 <- mean(train_ape_1 <= 0.15) * 100

# Absolute Percentage Error (APE) for test data
test_ape_1 <- abs(predictions_1 - test_data_1$sale_price) / test_data_1$sale_price

test_within_15_1 <- mean(test_ape_1 <= 0.15) * 100

print(c(train_within_15_1, test_within_15_1))

```


On the minimal model almost 31% of the training data is predicted within 15% of
the actual price, while in the test data 26%. This is rather low, so additional features
need to be included in the model.

### Adding sale year

```{r}
features_to_test <- model_df %>% 
  select(
    sale_price,
    longitude,
    latitude,
    sale_year,
    quality_score,
    bathrooms,
    bedrooms,
    has_pool_or_spa_detail
  ) %>% 
  drop_na() # in case, but here no NAs

cor_matrix <- cor(features_to_test)

price_correlations <- cor_matrix[, "sale_price"] %>% 
  sort(decreasing = TRUE)

price_correlations
```
We will first try to only add one feature, the one with the highest correlation to
sale price. This is the sale year.

```{r}
expanded_data <- model_df %>% 
  select(
    sale_price, 
    latitude, 
    longitude, 
    square_feet,
    sale_year # newly added feature
  ) %>% 
  drop_na() # none, but just in case


train_data_2 <- expanded_data[train_index, ]
test_data_2  <- expanded_data[-train_index, ]
```

```{r}

set.seed(222)
knn_v2 <- train(
  sale_price ~ latitude + longitude + square_feet + sale_year,
  data = train_data_2,
  method = "knn", 
  trControl = cv_control,
  preProcess = c("center", "scale"), 
  tuneGrid = k_grid
)

print(knn_v2)
```
```{r}
saveRDS(knn_v2, here::here("models", "knn_2.rds"))
```

By adding the sale year the optimal k dropped to 2, which is great for simplicity.
The RMSE is 153 705 a significant decrease from the
minimal model (247 381). The R-squared has also increased to 0.75.


```{r}
predictions_2 <- predict(knn_v2, newdata = test_data_2)

# Calculate performance metrics (RMSE, R-squared, MAE)
evaluation_metrics_2 <- postResample(pred = predictions_2, obs = test_data_2$sale_price)
print(evaluation_metrics_2)
```
The RMSE decrease and the R-squared increase is also consistent on the test set.

```{r}
test_ape_2 <- abs(predictions_2 - test_data_2$sale_price) / test_data_2$sale_price
test_within_15_2 <- mean(test_ape_2 <= 0.15) * 100

train_preds_2 <- predict(knn_v2, newdata = train_data_2)
train_ape_2 <- abs(train_preds_2 - train_data_2$sale_price) / train_data_2$sale_price
train_within_15_2 <- mean(train_ape_2 <= 0.15) * 100

print(c(train_within_15_2, test_within_15_2))
```

To our custom benchmark, by adding sale year to this model, it performs better on
the train set, namely almost 87% of the test data is predicted within 15% from the 
actual market value (previous model: 31%). Interestingly the performance on the 
training data is remarkably worse, 66.5% but still singificant improvement from the
previous 26%.

### Adding quality score

We add another variable with the second highest correlation to the target variable 
on an additional dimension (quality) next to location, size and time.

```{r}
expanded_data_2 <- model_df %>% 
  select(
    sale_price, 
    latitude, 
    longitude, 
    square_feet,
    sale_year,
    quality_score  # newly added feature
  ) %>% 
  drop_na() # none, but just in case

train_data_3 <- expanded_data_2[train_index, ]
test_data_3  <- expanded_data_2[-train_index, ]
```

```{r}

set.seed(222)
knn_v3 <- train(
  sale_price ~ latitude + longitude + square_feet + sale_year + quality_score,
  data = train_data_3,
  method = "knn", 
  trControl = cv_control,
  preProcess = c("center", "scale"), 
  tuneGrid = k_grid
)

print(knn_v3)
```
```{r}
saveRDS(knn_v3, here::here("models", "knn_3.rds"))
```

Again the optimal value for k is 2. RMSE showed a minor improvement to 140 185, compared
to previous 153 705 . The R-squared imporved from 0.74 to 0.79.


```{r}
predictions_3 <- predict(knn_v3, newdata = test_data_3)

# Calculate performance metrics (RMSE, R-squared, MAE)
evaluation_metrics_3 <- postResample(pred = predictions_3, obs = test_data_3$sale_price)
print(evaluation_metrics_3)
```

Adding the quality_score yields even better results on the test set, reducing the
RSME to 118 931, and increasing the R-squared above 80.

```{r}
test_ape_3 <- abs(predictions_3 - test_data_3$sale_price) / test_data_3$sale_price
test_within_15_3 <- mean(test_ape_3 <= 0.15) * 100

train_preds_3 <- predict(knn_v3, newdata = train_data_3)
train_ape_3 <- abs(train_preds_3 - train_data_3$sale_price) / train_data_3$sale_price
train_within_15_3 <- mean(train_ape_3 <= 0.15) * 100

print(c(train_within_15_3, test_within_15_3))
```
To our custom benchmark, by adding quality_score to this model, it performs marginally better on
the train set, namely  88.5% of the test data is predicted within 15% from the 
actual market value (previous model: 87%). Interestingly the performance on the 
training data show the same marginal improvement, almost 70% compared to previous 66.5%.

### All features

For the sake of seeing how much explanation or preciseness we are trading for better
user experience, we will fit a model that uses all selected predictors from above. 
Namely:
- **Physical Characteristics**: `square_feet`, `bathrooms`, `bedrooms`
- **Quality Indicators**: `quality_score`, `has_pool_or_spa_detail`
- **Location**: `latitude`, `longitude`
- **Time**: `sale year`.

```{r}
df_full <- model_df %>% 
  select(
    sale_price, 
    latitude, 
    longitude, 
    square_feet,
    sale_year,
    quality_score,
    bedrooms,
    bathrooms,
    has_pool_or_spa_detail
  ) %>% 
  drop_na() # none, but just in case

train_data_4 <- df_full[train_index, ]
test_data_4  <- df_full[-train_index, ]
```

```{r}
set.seed(222)
knn_full <- train(
  sale_price ~ latitude + longitude + square_feet + sale_year + quality_score + bathrooms
  + bedrooms,
  data = train_data_4,
  method = "knn", 
  trControl = cv_control,
  preProcess = c("center", "scale"), 
  tuneGrid = data.frame(k=k_grid)
)

print(knn_full)
```

```{r}
saveRDS(knn_full, here::here("models", "knn_full.rds"))
```

When all available features were used, the RMSE actually started to increase. At 
the same time the R-squared dropped slightly.

```{r}
predictions_full <- predict(knn_full, newdata = test_data_4)

# Calculate performance metrics (RMSE, R-squared, MAE)
evaluation_metrics_full <- postResample(pred = predictions_full, obs = test_data_4$sale_price)
print(evaluation_metrics_full)
```
The performance on the test data also slightly worsened with a 5 000 dollar increase
in RMSE compared to the best model.

```{r}
test_ape_full <- abs(predictions_full - test_data_4$sale_price) / test_data_4$sale_price
test_within_15_full <- mean(test_ape_full <= 0.15) * 100

train_preds_full <- predict(knn_full, newdata = train_data_4)
train_ape_full <- abs(train_preds_full - train_data_4$sale_price) / train_data_4$sale_price
train_within_15_full <- mean(train_ape_full <= 0.15) * 100

print(c(train_within_15_full, test_within_15_full))
```
To our custom benchmark, we see a slight drop in the percentage of observations predicted within 15%
of actual price in both train and test data.

## Step 3: Model Evaluation

| Model | RMSE | R-squared | MAE | 
|:----------|:-------|:-------|:-------|
| **minimal** | `r evaluation_metrics_1[["RMSE"]]` | `r evaluation_metrics_1[["Rsquared"]]` | `r evaluation_metrics_1[["MAE"]]` |
| **+sale_year** | `r evaluation_metrics_2[["RMSE"]]`  | `r evaluation_metrics_2[["Rsquared"]]` | `r evaluation_metrics_2[["MAE"]]` |
| **+sale_year+quality** | `r evaluation_metrics_3[["RMSE"]]`  | `r evaluation_metrics_3[["Rsquared"]]` | `r evaluation_metrics_3[["MAE"]]` |
| **full** | `r evaluation_metrics_full[["RMSE"]]` | `r evaluation_metrics_full[["Rsquared"]]` | `r evaluation_metrics_full[["MAE"]]` |


| Model | within 15% train | within 15% test
|:----------|:-------|:-------|
| **minimal** | `r train_within_15_1` | `r test_within_15_1` |
| **+sale_year** | `r train_within_15_2` | `r test_within_15_2` |
| **+sale_year+quality** | `r train_within_15_3` | `r test_within_15_3` |
| **full** | `r train_within_15_full` | `r test_within_15_full` |

```{r}
# assemble results
metrics_table <- data.frame(
  Model = c("minimal", "+sale_year", "+sale_year+quality", "full"),
  RMSE = c(evaluation_metrics_1[["RMSE"]], evaluation_metrics_2[["RMSE"]], 
           evaluation_metrics_3[["RMSE"]], evaluation_metrics_full[["RMSE"]]),
  Rsquared = c(evaluation_metrics_1[["Rsquared"]], evaluation_metrics_2[["Rsquared"]], 
               evaluation_metrics_3[["Rsquared"]], evaluation_metrics_full[["Rsquared"]]),
  MAE = c(evaluation_metrics_1[["MAE"]], evaluation_metrics_2[["MAE"]], 
          evaluation_metrics_3[["MAE"]], evaluation_metrics_full[["MAE"]])
)

within_15_table <- data.frame(
  Model = c("minimal", "+sale_year", "+sale_year+quality", "full"),
  within_15_train = c(train_within_15_1, train_within_15_2, train_within_15_3, train_within_15_full),
  within_15_test  = c(test_within_15_1,  test_within_15_2,  test_within_15_3,  test_within_15_full)
)

```

```{r}
# save
saveRDS(metrics_table,    here::here("analysis", "03-knn-files", "metrics_table.rds"))
saveRDS(within_15_table,  here::here("analysis", "03-knn-files", "within_15_table.rds"))
```

