---
title: "KNN model"
format:
  html:
    toc: true
    number-sections: true
---

# Introduction

This document provides a detailed interpretation of the KNN model applied to predict residential property prices in Pierce County. The analysis builds upon the data prepared during the exploratory phase and focuses on identifying the key drivers of house prices and assessing the predictive power of the resulting models.

# Step-by-Step Analysis

```{r}
#| label: setup
#| message: false
#| warning: false

library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(tidyr)
library(here)
library(dotenv)
library(caret)
library(tidyverse)
library(kknn)
options(scipen = 999)

# Load the prepared data
model_df <- read_csv(here::here("data", "house_price_model_data.csv"))
```


## Step 1: Data Preparation and Feature Selection

The analysis utilizes the `house_price_model_data.csv` dataset, which contains processed features including property characteristics and location data.

```{r}
#| label: data-summary
#| echo: false

cat("Total observations:", nrow(model_df), "\n")
cat("Total features available:", ncol(model_df) - 1, "\n")
```

The idea behind experimenting with KNN is to easily show the user why the given price
has been estimated through a clear reasoning (showing n most similar observations)
based on a very limited number of features (location and size, optionally condition).

Key features include:

- **Physical Characteristics**: `square_feet`, `bathrooms`, `bedrooms`
- **Quality Indicators**: `quality_score`, `has_pool_or_spa_detail`
- **Location**: `latitude`, `longitude`
- **Time**: `sale year`.

All models are fitted on log-transformed sale prices to reduce the influence of
extreme values and improve model fit. Evaluation metrics are back-transformed to
the original dollar scale for interpretability.

## Step 2: Model Development

### Minimal Model
We focus on the **Core Features Model** as it provides the most interpretable results
to the user with using only longitude, latitude and
the net square footage of properties.

```{r}
#| label: model-development

baseline_data <- model_df %>% 
  select(
    sale_price, 
    latitude, 
    longitude, 
    square_feet
  ) %>% 
  drop_na() %>%  # in case, but no NA
  mutate(log_sale_price = log(sale_price))

set.seed(222) 
train_index <- createDataPartition(
  baseline_data$log_sale_price, 
  p = 0.8, 
  list = FALSE
)

train_data_1 <- baseline_data[train_index, ]
test_data_1  <- baseline_data[-train_index, ]
```

We try for different values of k (2, 3, 4, 5). We decided these values as in a map
visualization (showing the closest observations to the user) more than 5 similar 
properties displayed would compromise the user experience.
```{r}

cv_control <- trainControl(
  method = "cv", 
  number = 3
)

k_grid <- data.frame(k = c(2, 3, 4, 5))

set.seed(222)
knn_baseline <- train(
  log_sale_price ~ latitude + longitude + square_feet,
  data = train_data_1,
  method = "knn",                  
  trControl = cv_control,
  preProcess = c("center", "scale"),
  tuneGrid = k_grid
)

print(knn_baseline)

```
```{r}
saveRDS(knn_baseline, here::here( "models", "knn_baseline.rds"))
```

The optimal number of k is 5. Predictions are made on the log scale and then
back-transformed via `exp()` for evaluation on the original dollar scale.

```{r}
# Predict on the test data (log scale), then back-transform
log_predictions_1 <- predict(knn_baseline, newdata = test_data_1)
predictions_1 <- exp(log_predictions_1)

# Helper: compute RMSE, R-squared, MAE on original dollar scale
dollar_metrics <- function(preds, actuals) {
  rmse <- sqrt(mean((preds - actuals)^2))
  ss_res <- sum((actuals - preds)^2)
  ss_tot <- sum((actuals - mean(actuals))^2)
  rsq <- 1 - ss_res / ss_tot
  mape <- mean(abs((actuals - preds) / actuals)) * 100
  c(RMSE = rmse, Rsquared = rsq, MAPE = mape)
}

# Calculate performance metrics on original dollar scale
evaluation_metrics_1 <- dollar_metrics(predictions_1, test_data_1$sale_price)
print(evaluation_metrics_1)

```


```{r}
log_train_preds_1 <- predict(knn_baseline, newdata = train_data_1)
train_preds_1 <- exp(log_train_preds_1)

# Absolute Percentage Error (APE) for training data
train_ape_1 <- abs(train_preds_1 - train_data_1$sale_price) / train_data_1$sale_price

train_within_15_1 <- mean(train_ape_1 <= 0.15) * 100

# Absolute Percentage Error (APE) for test data
test_ape_1 <- abs(predictions_1 - test_data_1$sale_price) / test_data_1$sale_price

test_mape_1 <- mean(test_ape_1) * 100

test_within_15_1 <- mean(test_ape_1 <= 0.15) * 100

print(c(train_within_15_1, test_within_15_1))

```


On the minimal model, the percentage of predictions within 15% of the actual price
is rather poor, which means additional features should be included.

### Adding sale year

```{r}
features_to_test <- model_df %>% 
  select(
    sale_price,
    longitude,
    latitude,
    sale_year,
    quality_score,
    bathrooms,
    bedrooms,
    has_pool_or_spa_detail
  ) %>% 
  drop_na() %>% # in case, but here no NAs
  mutate(log_sale_price = log(sale_price))

cor_matrix <- cor(features_to_test)

price_correlations <- cor_matrix[, "log_sale_price"] %>% 
  sort(decreasing = TRUE)

price_correlations
```
We will first try to only add one feature, the one with the highest correlation to
log sale price. This is the sale year.

```{r}
expanded_data <- model_df %>% 
  select(
    sale_price, 
    latitude, 
    longitude, 
    square_feet,
    sale_year # newly added feature
  ) %>% 
  drop_na() %>% # none, but just in case
  mutate(log_sale_price = log(sale_price))


train_data_2 <- expanded_data[train_index, ]
test_data_2  <- expanded_data[-train_index, ]
```

```{r}

set.seed(222)
knn_v2 <- train(
  log_sale_price ~ latitude + longitude + square_feet + sale_year,
  data = train_data_2,
  method = "knn", 
  trControl = cv_control,
  preProcess = c("center", "scale"), 
  tuneGrid = k_grid
)

print(knn_v2)
```
```{r}
saveRDS(knn_v2, here::here("models", "knn_2.rds"))
```

By adding the sale year the optimal k remains at 5.
The cross-validation RMSE on the log scale improved significantly.

```{r}
log_predictions_2 <- predict(knn_v2, newdata = test_data_2)
predictions_2 <- exp(log_predictions_2)

# Calculate performance metrics on original dollar scale
evaluation_metrics_2 <- dollar_metrics(predictions_2, test_data_2$sale_price)
print(evaluation_metrics_2)
```
The RMSE decrease and the R-squared increase on the dollar scale is consistent with
the cross-validation results.

```{r}
log_test_preds_2 <- predict(knn_v2, newdata = test_data_2)
test_ape_2 <- abs(exp(log_test_preds_2) - test_data_2$sale_price) / test_data_2$sale_price
test_within_15_2 <- mean(test_ape_2 <= 0.15) * 100

log_train_preds_2 <- predict(knn_v2, newdata = train_data_2)
train_preds_2 <- exp(log_train_preds_2)
train_ape_2 <- abs(train_preds_2 - train_data_2$sale_price) / train_data_2$sale_price
train_within_15_2 <- mean(train_ape_2 <= 0.15) * 100

print(c(train_within_15_2, test_within_15_2))
```

By adding sale year, the share of predictions within 15% of actual market value
improves substantially on both train and test sets compared to the minimal model.

### Adding quality score

We add another variable with the second highest correlation to the log target variable 
on an additional dimension (quality) next to location, size and time.

```{r}
expanded_data_2 <- model_df %>% 
  select(
    sale_price, 
    latitude, 
    longitude, 
    square_feet,
    sale_year,
    quality_score  # newly added feature
  ) %>% 
  drop_na() %>% # none, but just in case
  mutate(log_sale_price = log(sale_price))

train_data_3 <- expanded_data_2[train_index, ]
test_data_3  <- expanded_data_2[-train_index, ]
```

```{r}

set.seed(222)
knn_v3 <- train(
  log_sale_price ~ latitude + longitude + square_feet + sale_year + quality_score,
  data = train_data_3,
  method = "knn", 
  trControl = cv_control,
  preProcess = c("center", "scale"), 
  tuneGrid = k_grid
)

print(knn_v3)
```
```{r}
saveRDS(knn_v3, here::here("models", "knn_3.rds"))
```

Again the optimal value for k is 5. The cross-validation RMSE on the log scale
showed a further improvement compared to the previous model.

```{r}
log_predictions_3 <- predict(knn_v3, newdata = test_data_3)
predictions_3 <- exp(log_predictions_3)

# Calculate performance metrics on original dollar scale
evaluation_metrics_3 <- dollar_metrics(predictions_3, test_data_3$sale_price)
print(evaluation_metrics_3)
```

Adding the quality_score yields even better results on the test set, further
reducing the RMSE and increasing the R-squared on the original dollar scale.

```{r}
test_ape_3 <- abs(predictions_3 - test_data_3$sale_price) / test_data_3$sale_price
test_within_15_3 <- mean(test_ape_3 <= 0.15) * 100

log_train_preds_3 <- predict(knn_v3, newdata = train_data_3)
train_preds_3 <- exp(log_train_preds_3)
train_ape_3 <- abs(train_preds_3 - train_data_3$sale_price) / train_data_3$sale_price
train_within_15_3 <- mean(train_ape_3 <= 0.15) * 100

print(c(train_within_15_3, test_within_15_3))
```
Adding quality_score leads to a marginal further improvement in the share of 
predictions within 15% of actual market value on both train and test sets.

### All features

For the sake of seeing how much explanation or preciseness we are trading for better
user experience, we will fit a model that uses all selected predictors from above. 
Namely:

- **Physical Characteristics**: `square_feet`, `bathrooms`, `bedrooms`
- **Quality Indicators**: `quality_score`, `has_pool_or_spa_detail`
- **Location**: `latitude`, `longitude`
- **Time**: `sale year`.

```{r}
df_full <- model_df %>% 
  select(
    sale_price, 
    latitude, 
    longitude, 
    square_feet,
    sale_year,
    quality_score,
    bedrooms,
    bathrooms,
    has_pool_or_spa_detail
  ) %>% 
  drop_na() %>% # none, but just in case
  mutate(log_sale_price = log(sale_price))

train_data_4 <- df_full[train_index, ]
test_data_4  <- df_full[-train_index, ]
```

```{r}
set.seed(222)
knn_full <- train(
  log_sale_price ~ latitude + longitude + square_feet + sale_year + quality_score + bathrooms
  + bedrooms,
  data = train_data_4,
  method = "knn", 
  trControl = cv_control,
  preProcess = c("center", "scale"), 
  tuneGrid = data.frame(k=k_grid)
)

print(knn_full)
```

```{r}
saveRDS(knn_full, here::here("models", "knn_full.rds"))
```

When all available features were used, the cross-validation RMSE on the log scale
started to increase, suggesting the curse of dimensionality may be kicking in. The
R-squared also dropped slightly.

```{r}
log_predictions_full <- predict(knn_full, newdata = test_data_4)
predictions_full <- exp(log_predictions_full)

# Calculate performance metrics on original dollar scale
evaluation_metrics_full <- dollar_metrics(predictions_full, test_data_4$sale_price)
print(evaluation_metrics_full)
```
The dollar-scale performance on the test set also slightly worsened compared to 
the best model.

```{r}
test_ape_full <- abs(predictions_full - test_data_4$sale_price) / test_data_4$sale_price
test_within_15_full <- mean(test_ape_full <= 0.15) * 100

log_train_preds_full <- predict(knn_full, newdata = train_data_4)
train_preds_full <- exp(log_train_preds_full)
train_ape_full <- abs(train_preds_full - train_data_4$sale_price) / train_data_4$sale_price
train_within_15_full <- mean(train_ape_full <= 0.15) * 100

print(c(train_within_15_full, test_within_15_full))
```
A slight drop in the percentage of observations predicted within 15% of actual 
price is observed in both train and test data, confirming the full model does not
outperform the more parsimonious one.

## Step 3: Model Evaluation

All metrics below are computed on the **original dollar scale** (after back-transforming
log predictions via `exp()`).

| Model | RMSE | R-squared | MAE | 
|:----------|:-------|:-------|:-------|
| **minimal** | `r evaluation_metrics_1[["RMSE"]]` | `r evaluation_metrics_1[["Rsquared"]]` | `r evaluation_metrics_1[["MAPE"]]` |
| **+sale_year** | `r evaluation_metrics_2[["RMSE"]]`  | `r evaluation_metrics_2[["Rsquared"]]` | `r evaluation_metrics_2[["MAPE"]]` |
| **+sale_year+quality** | `r evaluation_metrics_3[["RMSE"]]`  | `r evaluation_metrics_3[["Rsquared"]]` | `r evaluation_metrics_3[["MAPE"]]` |
| **full** | `r evaluation_metrics_full[["RMSE"]]` | `r evaluation_metrics_full[["Rsquared"]]` | `r evaluation_metrics_full[["MAPE"]]` |


| Model | within 15% train | within 15% test
|:----------|:-------|:-------|
| **minimal** | `r train_within_15_1` | `r test_within_15_1` |
| **+sale_year** | `r train_within_15_2` | `r test_within_15_2` |
| **+sale_year+quality** | `r train_within_15_3` | `r test_within_15_3` |
| **full** | `r train_within_15_full` | `r test_within_15_full` |

```{r}
# assemble results
metrics_table <- data.frame(
  Model = c("minimal", "+sale_year", "+sale_year+quality", "full"),
  RMSE = c(evaluation_metrics_1[["RMSE"]], evaluation_metrics_2[["RMSE"]], 
           evaluation_metrics_3[["RMSE"]], evaluation_metrics_full[["RMSE"]]),
  Rsquared = c(evaluation_metrics_1[["Rsquared"]], evaluation_metrics_2[["Rsquared"]], 
               evaluation_metrics_3[["Rsquared"]], evaluation_metrics_full[["Rsquared"]]),
  MAPE = c(evaluation_metrics_1[["MAPE"]], evaluation_metrics_2[["MAPE"]], 
          evaluation_metrics_3[["MAPE"]], evaluation_metrics_full[["MAPE"]])
)

within_15_table <- data.frame(
  Model = c("minimal", "+sale_year", "+sale_year+quality", "full"),
  within_15_train = c(train_within_15_1, train_within_15_2, train_within_15_3, train_within_15_full),
  within_15_test  = c(test_within_15_1,  test_within_15_2,  test_within_15_3,  test_within_15_full)
)

```

```{r}
metrics_table
```
```{r}
within_15_table
```


```{r}
# save
saveRDS(metrics_table,    here::here("analysis", "03-knn-files", "metrics_table.rds"))
#saveRDS(within_15_table,  here::here("analysis", "03-knn-files", "within_15_table.rds"))
```
