---
title: "LightGBM"
format: html
editor: visual
---

```{r}
library(dplyr)
library(tidyr)
library(caret)
library(lightgbm)
library(ggplot2)

```

# LightGBM - gradient boosting algorithm

LightGBM is a gradient boosting algorithm that builds many decision trees sequentially, where each new tree tries to correct the errors made by the previous trees. It is efficient for large tabular datasets and can capture non-linear relationships and feature interactions well.

## Prepare LightGBM dataframe

```{r}

app_model_df <- read.csv(
  "/workspaces/SOE-R-House-Price-Prediction/data/house_price_model_data.csv",
  header = TRUE,
  stringsAsFactors = FALSE
)

# Select app-focused model features
app_model_features <- c(
  "sale_price",
  "latitude",
  "longitude",
  "location_cluster",
  "sale_year",
  "square_feet",
  "land_net_square_feet",
  "porch_square_feet",
  "attached_garage_square_feet",
  "quality_score",
  "bedrooms",
  "bathrooms",
  "bedrooms_per_1000_sqft",
  "bathrooms_per_1000_sqft",
  "year_built",
  "has_deck_or_porch_detail",
  "has_garage_or_carport_detail",
  "has_pool_or_spa_detail",
  "has_waterfront_detail",
  "has_finished_space_detail"
)

# Prepare LightGBM dataframe
lgb_df <- app_model_df %>%
  select(any_of(app_model_features)) %>%
  distinct() %>%
  drop_na()

# Convert target and categorical/boolean variables
lgb_df <- lgb_df %>%
  mutate(
    log_sale_price = log(sale_price),
    location_cluster = as.factor(location_cluster),
    across(where(is.character), as.factor),
    across(where(is.logical), as.factor)
  )

head(lgb_df)
```

## Train-Test Split and Data Preparation for LightGBM

```{r}
# Train-test split
set.seed(123)

train_index <- createDataPartition(
  lgb_df$log_sale_price,
  p = 0.8,
  list = FALSE
)

lgb_train <- lgb_df[train_index, ]
lgb_test <- lgb_df[-train_index, ]

# Convert categorical variables into dummy variables
dummy_model_lgb <- dummyVars(
  log_sale_price ~ . - sale_price,
  data = lgb_train,
  fullRank = TRUE
)

x_train_lgb <- predict(dummy_model_lgb, newdata = lgb_train) %>%
  as.data.frame()

x_test_lgb <- predict(dummy_model_lgb, newdata = lgb_test) %>%
  as.data.frame()

# Target variables
y_train_lgb <- lgb_train$log_sale_price
y_test_lgb <- lgb_test$log_sale_price

# Convert to LightGBM dataset format
dtrain_lgb <- lgb.Dataset(
  data = as.matrix(x_train_lgb),
  label = y_train_lgb
)

dtest_lgb <- lgb.Dataset(
  data = as.matrix(x_test_lgb),
  label = y_test_lgb
)

```

## Train LightGBM Model

```{r}
# Set LightGBM parameters
lgb_params <- list(
  objective = "regression",
  metric = "rmse",
  learning_rate = 0.05,
  num_leaves = 31,
  max_depth = -1,
  feature_fraction = 0.8,
  bagging_fraction = 0.8,
  bagging_freq = 5,
  min_data_in_leaf = 20
)

# Train LightGBM model
set.seed(123)

lgb_model <- lgb.train(
  params = lgb_params,
  data = dtrain_lgb,
  nrounds = 1000,
  valids = list(
    train = dtrain_lgb,
    test = dtest_lgb
  ),
  early_stopping_rounds = 50,
  verbose = 50
)

lgb_model

# Predict on test data
lgb_pred_log <- predict(
  lgb_model,
  newdata = as.matrix(x_test_lgb)
)

# Convert predictions back to USD scale
lgb_results <- data.frame(
  actual_log_sale_price = y_test_lgb,
  predicted_log_sale_price = lgb_pred_log
) %>%
  mutate(
    actual_sale_price = exp(actual_log_sale_price),
    predicted_sale_price = exp(predicted_log_sale_price),
    percent_error = abs(predicted_sale_price - actual_sale_price) / actual_sale_price,
    within_15_percent = percent_error <= 0.15
  )

head(lgb_results)

```

## Evaluate Model Performance and Analyze Results

```{r}
# Evaluate model
lgb_metrics <- lgb_results %>%
  summarise(
    RMSE = sqrt(mean((actual_sale_price - predicted_sale_price)^2)),
    MAE = mean(abs(actual_sale_price - predicted_sale_price)),
    R_squared = cor(actual_sale_price, predicted_sale_price)^2,
    MAPE = mean(percent_error) * 100,
    percent_within_15 = mean(within_15_percent) * 100
  )

lgb_metrics

# Actual vs predicted plot
ggplot(lgb_results, aes(x = actual_sale_price, y = predicted_sale_price)) +
  geom_point(alpha = 0.3) +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "Actual vs Predicted Sale Price - LightGBM",
    x = "Actual Sale Price in USD, Log Scale",
    y = "Predicted Sale Price in USD, Log Scale"
  )

# Prediction error distribution
ggplot(lgb_results, aes(x = percent_error)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Distribution of LightGBM Prediction Error",
    x = "Absolute Percentage Error",
    y = "Number of Properties"
  )

```

## Feature Importance and Model Saving

```{r}
# Feature importance
lgb_importance <- lgb.importance(
  lgb_model,
  percentage = TRUE
)

head(lgb_importance, 10)

lgb.plot.importance(
  lgb_importance,
  top_n = 10
)

# Save LightGBM model and dummy transformation object
lgb.save(
  lgb_model,
  "lightgbm_house_price_model.txt"
)

saveRDS(
  dummy_model_lgb,
  "lightgbm_dummy_model.rds"
)
```

```{r}
# -------------------------------
# Tune LightGBM Model
# -------------------------------

library(dplyr)
library(lightgbm)

# Define tuning grid
lgb_grid <- expand.grid(
  learning_rate = c(0.03, 0.05, 0.08),
  num_leaves = c(31, 63),
  max_depth = c(-1, 8),
  min_data_in_leaf = c(20, 50),
  feature_fraction = c(0.8),
  bagging_fraction = c(0.8)
)

# Empty table to store results
lgb_tuning_results <- data.frame()

# Loop through all parameter combinations
for (i in 1:nrow(lgb_grid)) {
  
  params <- list(
    objective = "regression",
    metric = "rmse",
    learning_rate = lgb_grid$learning_rate[i],
    num_leaves = lgb_grid$num_leaves[i],
    max_depth = lgb_grid$max_depth[i],
    min_data_in_leaf = lgb_grid$min_data_in_leaf[i],
    feature_fraction = lgb_grid$feature_fraction[i],
    bagging_fraction = lgb_grid$bagging_fraction[i],
    bagging_freq = 5
  )
  
  set.seed(123)
  
  lgb_model_tuned <- lgb.train(
    params = params,
    data = dtrain_lgb,
    nrounds = 1000,
    valids = list(test = dtest_lgb),
    early_stopping_rounds = 50,
    verbose = -1
  )
  
  # Predict on test data
  lgb_pred_log <- predict(
    lgb_model_tuned,
    newdata = as.matrix(x_test_lgb)
  )
  
  # Convert predictions back to USD scale
  lgb_results_tuned <- data.frame(
    actual_log_sale_price = y_test_lgb,
    predicted_log_sale_price = lgb_pred_log
  ) %>%
    mutate(
      actual_sale_price = exp(actual_log_sale_price),
      predicted_sale_price = exp(predicted_log_sale_price),
      percent_error = abs(predicted_sale_price - actual_sale_price) / actual_sale_price,
      within_15_percent = percent_error <= 0.15
    )
  
  # Calculate metrics
  metrics <- lgb_results_tuned %>%
    summarise(
      learning_rate = lgb_grid$learning_rate[i],
      num_leaves = lgb_grid$num_leaves[i],
      max_depth = lgb_grid$max_depth[i],
      min_data_in_leaf = lgb_grid$min_data_in_leaf[i],
      feature_fraction = lgb_grid$feature_fraction[i],
      bagging_fraction = lgb_grid$bagging_fraction[i],
      best_iter = lgb_model_tuned$best_iter,
      RMSE = sqrt(mean((actual_sale_price - predicted_sale_price)^2)),
      MAE = mean(abs(actual_sale_price - predicted_sale_price)),
      R_squared = cor(actual_sale_price, predicted_sale_price)^2,
      MAPE = mean(percent_error) * 100,
      percent_within_15 = mean(within_15_percent) * 100
    )
  
  # Store metrics
  lgb_tuning_results <- bind_rows(
    lgb_tuning_results,
    metrics
  )
}

# View tuning results sorted by RMSE
lgb_tuning_results %>%
  arrange(RMSE)
```

```{r}
lgb_tuning_results %>%
  arrange(desc(percent_within_15))
```

The tuning results show that the best LightGBM model used a learning rate of `0.08`, `63` leaves, a maximum depth of `8`, and `20` minimum observations per leaf. This model achieved the lowest RMSE of 83,321 USD and an R-squared of 0.924, meaning it explains about 92.4 percent of the variation in sale price.

Overall, models with `num_leaves = 63` performed better than models with `num_leaves = 31`, suggesting that the model benefited from more complex tree structures. The best results were also achieved with `learning_rate = 0.08`, meaning this learning rate worked better than the smaller values tested in this tuning grid.

# Tuned model with most important features

After tuning the LightGBM model, a reduced feature set based on feature importance can be tested. The goal is to check whether the model can achieve similar performance with fewer input variables. This is especially important for the final app, because a simpler input form improves usability. However, features should only be removed if model performance does not decrease meaningfully.

```{r}
library(dplyr)
library(tidyr)
library(caret)
library(lightgbm)

# Features ordered from MOST important to LEAST important
feature_importance_order <- c(
  "sale_year",
  "quality_score",
  "year_built",
  "square_feet",
  "latitude",
  "land_net_square_feet",
  "bathrooms",
  "longitude",
  "bedrooms_per_1000_sqft",
  "bathrooms_per_1000_sqft",
  "has_waterfront_detail",
  "location_cluster",
  "bedrooms",
  "has_finished_space_detail",
  "has_garage_or_carport_detail"
)

all_lgb_features <- c(
  "sale_price",
  feature_importance_order
)

# Prepare base dataset
lgb_base_df <- app_model_df %>%
  select(any_of(all_lgb_features)) %>%
  distinct() %>%
  drop_na() %>%
  mutate(
    log_sale_price = log(sale_price),
    across(where(is.character), as.factor),
    across(where(is.logical), as.factor),
    location_cluster = as.factor(location_cluster)
  )

# Same train-test split for every model
set.seed(123)

train_index <- createDataPartition(
  lgb_base_df$log_sale_price,
  p = 0.8,
  list = FALSE
)

lgb_train_base <- lgb_base_df[train_index, ]
lgb_test_base <- lgb_base_df[-train_index, ]

# Best tuned LightGBM parameters
best_lgb_params <- list(
  objective = "regression",
  metric = "rmse",
  learning_rate = 0.08,
  num_leaves = 63,
  max_depth = 8,
  min_data_in_leaf = 20,
  feature_fraction = 0.8,
  bagging_fraction = 0.8,
  bagging_freq = 5
)

# Function to train and evaluate one model
train_evaluate_lgb_stepwise <- function(features_to_use, removed_features) {
  
  train_data <- lgb_train_base %>%
    select(all_of(c("sale_price", "log_sale_price", features_to_use)))
  
  test_data <- lgb_test_base %>%
    select(all_of(c("sale_price", "log_sale_price", features_to_use)))
  
  dummy_model <- dummyVars(
    log_sale_price ~ . - sale_price,
    data = train_data,
    fullRank = TRUE
  )
  
  x_train <- predict(dummy_model, newdata = train_data) %>%
    as.data.frame()
  
  x_test <- predict(dummy_model, newdata = test_data) %>%
    as.data.frame()
  
  y_train <- train_data$log_sale_price
  y_test <- test_data$log_sale_price
  
  dtrain <- lgb.Dataset(
    data = as.matrix(x_train),
    label = y_train
  )
  
  dtest <- lgb.Dataset(
    data = as.matrix(x_test),
    label = y_test
  )
  
  set.seed(123)
  
  model <- lgb.train(
    params = best_lgb_params,
    data = dtrain,
    nrounds = 1000,
    valids = list(test = dtest),
    early_stopping_rounds = 50,
    verbose = -1
  )
  
  pred_log <- predict(
    model,
    newdata = as.matrix(x_test)
  )
  
  results <- data.frame(
    actual_log_sale_price = y_test,
    predicted_log_sale_price = pred_log
  ) %>%
    mutate(
      actual_sale_price = exp(actual_log_sale_price),
      predicted_sale_price = exp(predicted_log_sale_price),
      percent_error = abs(predicted_sale_price - actual_sale_price) / actual_sale_price,
      within_15_percent = percent_error <= 0.15
    )
  
  metrics <- results %>%
    summarise(
      number_of_features_used = length(features_to_use),
      number_of_features_removed = length(removed_features),
      removed_features = paste(removed_features, collapse = ", "),
      features_used = paste(features_to_use, collapse = ", "),
      RMSE = sqrt(mean((actual_sale_price - predicted_sale_price)^2)),
      MAE = mean(abs(actual_sale_price - predicted_sale_price)),
      R_squared = cor(actual_sale_price, predicted_sale_price)^2,
      MAPE = mean(percent_error) * 100,
      percent_within_15 = mean(within_15_percent) * 100
    )
  
  return(metrics)
}

# Run stepwise removal:
# remove 1 least important, then 2 least important, then 3, etc.
stepwise_results <- data.frame()

for (n_remove in 1:(length(feature_importance_order) - 3)) {
  
  removed_features <- tail(feature_importance_order, n_remove)
  features_to_use <- setdiff(feature_importance_order, removed_features)
  
  cat(
    "Training model removing",
    n_remove,
    "least important feature(s):",
    paste(removed_features, collapse = ", "),
    "\n"
  )
  
  model_metrics <- train_evaluate_lgb_stepwise(
    features_to_use = features_to_use,
    removed_features = removed_features
  )
  
  stepwise_results <- bind_rows(
    stepwise_results,
    model_metrics
  )
}

stepwise_results_summary <- stepwise_results %>%
  select(
    number_of_features_removed,
    number_of_features_used,
    removed_features,
    RMSE,
    MAE,
    R_squared,
    MAPE,
    percent_within_15
  ) %>%
  mutate(
    RMSE = round(RMSE, 2),
    MAE = round(MAE, 2),
    R_squared = round(R_squared, 4),
    MAPE = round(MAPE, 2),
    percent_within_15 = round(percent_within_15, 2)
  ) %>%
  arrange(desc(percent_within_15))

print(stepwise_results_summary, width = Inf)

```

# Summary

```{r}
stepwise_results_summary_rmse <- stepwise_results %>%
  select(
    number_of_features_removed,
    number_of_features_used,
    removed_features,
    RMSE,
    MAE,
    R_squared,
    MAPE,
    percent_within_15
  ) %>%
  mutate(
    RMSE = round(RMSE, 2),
    MAE = round(MAE, 2),
    R_squared = round(R_squared, 4),
    MAPE = round(MAPE, 2),
    percent_within_15 = round(percent_within_15, 2)
  ) %>%
  arrange(RMSE)

print(stepwise_results_summary_rmse, width = Inf)
```

Two LightGBM feature-reduction models were compared to decide which model is more suitable for the final app.

| Model | Removed features | RMSE | MAE | R-squared | MAPE | Within 15% |
|-----------|-----------|----------:|----------:|----------:|----------:|----------:|
| Model that eliminates 3 features | `bedrooms`, `has_finished_space_detail`, `has_garage_or_carport_detail` | 88,896.91 | 41,668.10 | 0.9095 | 12.53% | 76.86% |
| Model that eliminates 7 features | `bedrooms_per_1000_sqft`, `bathrooms_per_1000_sqft`, `has_waterfront_detail`, `location_cluster`, `bedrooms`, `has_finished_space_detail`, `has_garage_or_carport_detail` | 90,574.49 | 42,210.63 | 0.9058 | 12.63% | 76.38% |
| Differences between models | Model with 7 removed features minus model with 3 removed features | +1,677.58 | +542.53 | -0.0037 | +0.10 % | -0.48 % |

The model that eliminates 7 features has slightly weaker performance than the model that eliminates 3 features, but the difference is small. RMSE increases by 1,677.58 USD, MAE increases by 542.53 USD, MAPE increases by only 0.10 percentage points, and the percentage of predictions within 15 percent decreases by only 0.48 percentage points.

Because the model that eliminates 7 features removes 4 additional input variables while keeping very similar performance, it may be more suitable for the final app from a usability perspective.

## LightGBM Conclusion

The final LightGBM model recommends using the following features: `sale_year`, `quality_score`, `year_built`, `square_feet`, `latitude`, `land_net_square_feet`, `bathrooms`, and `longitude`.

It suggests excluding `bedrooms_per_1000_sqft`, `bathrooms_per_1000_sqft`, `has_waterfront_detail`, `location_cluster`, `bedrooms`, `has_finished_space_detail`, and `has_garage_or_carport_detail`.